---
title: "Downscaling guidelines"
author: "Clara Douzal"
date: "2026-01-23"
output: pdf_document
header-includes:
  - \DeclareUnicodeCharacter{2061}{}
  - \DeclareUnicodeCharacter{2212}{}
params:
  country : "IND"
  pathway : "CT_BxG"
  start_map_source : "HILDA"
  mnl_niter : 100
  mnl_nburn : 50
  seed : 1212
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 150)

```



```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
library(stringr)
library(knitr)
library(readr)
#library(forcats)
#library(purrr)
#library(writexl)
library(ggplot2)

if (requireNamespace("conflicted", quietly = TRUE)) {
  conflicted::conflict_prefer("filter",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("select",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("lag",       "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("rename",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("left_join", "dplyr", quiet = TRUE)
}
```

```{r setting paths, include = FALSE}
set.seed(params$seed)
country <- params$country
pathway <- params$pathway
niter   <- params$mnl_niter
nburn   <- params$mnl_nburn
start_map_source <- params$start_map_source

dir.create(here("Output", country), showWarnings = FALSE, recursive = TRUE)
stamp <- format(Sys.Date(), "%y%m%d")
tag   <- paste(stamp, country, pathway, start_map_source, sep = "_")

```

This report is generated for **`r country`** for the pathway **`r pathway`**.

```{r loading special functions, include = FALSE}
## functions to load all the country specific data downloaded from Earth engine
source(here("Program", "functions", "LoadData.R"))


```

# Checking input data

```{r data load, include=FALSE, cache = FALSE}

# Spatial data used to compute priors
# datasets name that will be used in the R script and their associated geojson file names
spatialdatafiles <- rbind.data.frame(
  c("population" ,      "Population2020.geojson"),
  c("protectedareas" ,  "ProtectedAreas.geojson"),
  c("travel",           "TravelTime.geojson"),
  c("landcoverHILDA",        "LandCoverHILDA2015.geojson"),
  c("landcoverCopernicus",        "LandCoverCopernicus2019.geojson"),
  c("landcoverchange",  "LandCoverChangeHILDA2015_2019.geojson"),
  c("forestmanagement", "ForestManagement.geojson"),
  c("altitude",         "Altitude.geojson"),
  c("slope",            "Slope.geojson"),
  c("gaez",             "GAEZCropDistribution2015.geojson"),
  c("livestock",        "GLW4WorldGriddedLivestock2020.geojson"),
  c("ab_be_biomass",    "AbovegroundBelowgroundBiomassCarbonDensity.geojson")
  ) %>% 
  setNames(c("rdataset", "filename"))

#load spatial data
spatialdataset <- LoadSpatialData(country, spatialdatafiles)

if(start_map_source == "HILDA"){
spatialdataset$landcover <- spatialdataset$landcoverHILDA #this is in km2 
} else {
  spatialdataset$landcover <- spatialdataset$landcoverCopernicus
} 

# datasets name that will be used in the R script and their associated sheet names in mapping_code.xlsx
mappingdatafiles <- rbind.data.frame(
  c("map_HILDA" ,         "HILDA"),
  c("map_HILDA_LUC" ,     "HILDA_change"),
  c("map_Copernicus",     "Copernicus"),
  c("map_ForestMgmt",     "ForestManagement"),
  c("map_protectedareas", "ProtectedAreas")
  ) %>% 
  setNames(c("rdataset", "sheetname"))

mappingdataset <- LoadMappingData("mapping_code", mappingdatafiles)

# Spatial restoration datasets used to set constraints and spatial targets in the FABLE-C
# datasets name that will be used in the R script and their associated geojson file names


#### Should I read the restoration data here ??? ####
#### Maybe create a separate function that deals with this ####
```

<!-- Below is the table of input data used in the prior module of the downscaling process. -->
<!-- DON'T KNOW IF I SHOULD KEEP IT -->
```{r echo=FALSE, message=FALSE}


# # Create the data frame for the table
# table_data <- data.frame(
#   `Variable in econometric model` = c("Dependent", "Land use change", "", 
#                                       "Explanatory variables", "Land use", 
#                                       "Forest Management", "Mean Travel time to major cities", "Population", "Crop yield (27 indicidual crops)"),
#   Year = c("", "2015-2019", "", "", "2015", "2015", "2000", "2020", "2015"),
#   Unit = c("", "% of lu.from in pixel being converted", "", "", 
#            "hectares", "hectares", "minutes", "Count", "kg/ha"),
#   Source = c("", "HILDA", "", "", "COPERNICUS", "Lesiv et al.", 
#              "Nelson, A (2008)", "WorldPop", "GAEZ+")
# )
# 
# # Use kable without booktabs
# kable(
#   table_data, 
#   format = "latex", 
#   col.names = c("Variable in econometric model", "Year", "Unit", "Source")
# )
```

## Checking the land cover change input data

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix. With "missing_from" you can see the number of land transitions thatare missing because of a missing initial land cover. With "missing_to" you can see the number of land transitions that are missing because of a missing final land cover.


```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
#value is the percentage of the cell that is going from land cover to another
country_luc <- LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA_LUC %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water", "not relevant"))) %>% 
  filter(!(to %in% c("ocean", "water", "not relevant"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2015) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2015,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) 

# table(country_luc$lu.from, country_luc$lu.to)
  
expected <- c("cropland","forest","newforest","otherland","pasture","urban")
present_from <- sort(unique(country_luc$lu.from))
present_to   <- sort(unique(country_luc$lu.to))
data.frame(kind = c("missing_from","missing_to"),
           value = I(list(setdiff(expected, present_from),
                         setdiff(expected, present_to)))) %>% knitr::kable()
```