---
title: "Downscaling"
author: "Clara Douzal"
date: "2023-10-24"
output: pdf_document
params:
  variable1 : country
  variable2 : pathway 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
library(stringr)
```


```{r temporary settings, include = FALSE}
#Please ensure that you have a folder named like the country variable you will input below in your "Output" folder. This is where your maps will be saved.

country = "IND"
pathway = "CurrentTrends"
# pathway = "NationalCommitments"
# pathway = "GlobalSustainability"

```

This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population <- geojson_read(here("Data", country, "Population2020.geojson"),  what = "sp")
population.df <- population@data %>% 
  select(-id) #%>% 
  #mutate(TotalPop = c(scale(TotalPop, scale = T)))

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"),  what = "sp")
ProtectedAreas.df <- ProtectedAreas@data %>% 
  select(-id)

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"),  what = "sp")
Travel.df <- Travel@data %>% 
  select(-id)

#this is in km2 
LandCover <- geojson_read(here("Data", country, "LandCoverCopernicus2019.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)

#this is in km2
LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2018_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data %>% 
   select(-id)

ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data %>% 
  select(-id)

altitude <- geojson_read(here("Data", country, "Altitude.geojson"),  what = "sp")
altitude.df <- altitude@data %>% 
  select(-id)

slope <- geojson_read(here("Data", country, "Slope.geojson"),  what = "sp")
slope.df <- slope@data %>% 
  select(-id)
  

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")

map_ForestMgmt <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ForestManagement")
```

```{r create raster for maps, include = FALSE}
#raster to use for plots
ext <- extent(Travel)
res <- 0.5
raster_layer <- raster(ext, res = res)
rasterized_layer <- rasterize(Travel, raster_layer, field = "id")
```

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
#value is the percentage of the cell that is going from land cover to another
country_luc <- LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2018) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2018,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2018,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2018,
                   value = 0)) 

table(country_luc$lu.from, country_luc$lu.to)
  
```

```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}

Xmat_LU <- LandCover.df %>%
  # Pivot the land cover data to long format
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_Copernicus to get LandCover names
  left_join(map_Copernicus %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  # Summarize total area per land cover class
  group_by(id_c, LandCover) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha to match FABLE unit
  summarise(value = sum(AreaPerCover, na.rm = TRUE)/10, .groups = 'drop') %>%
  # Pivot wider to spread land cover types into columns
  pivot_wider(names_from = "LandCover", values_from = "value", values_fill = 0) #%>%
  
  ### This is the part where I compute the share that I will temporarily remove from the script while Tamas temas works on the updating of downscalR to include the dynamic update of the shares in Xmat after downscaling
  
  # # Remove unwanted land cover types
  # # select(-c(unknwon, water, ocean)) %>%
  # # Compute the share of each land use class
  # rowwise() %>%
  # mutate(
  #   notrelevant = sum(c_across(c(unknown, water, ocean, notrelevant)), na.rm = TRUE)) %>% 
  # select(-c(unknown, water, ocean)) %>%
  # mutate(
  #   # Calculate total area of selected land cover classes, avoid NaN by setting total to 1 if sum is 0
  #   total = ifelse(sum(c_across(c(cropland, forest, otherland, pasture, urban, notrelevant)), na.rm = TRUE) == 0, 
  #                  1, 
  #                  sum(c_across(c(cropland, forest, otherland, pasture, urban, notrelevant)), na.rm = TRUE)),
  #   # Compute shares for each land cover class
  #   across(c(cropland, forest, otherland, pasture, urban), ~ .x / total, .names = "{.col}_share")
  # ) %>%
  # ungroup() %>%
  # # Remove the total column because it won't be used
  # # Remove the notrelevant column to avoid any collinearity
  # select(-c(total, notrelevant))

#ForestManagement
Xmat_ForestManagement <- ForestManagement.df %>%
  # Pivot longer to convert cover columns to rows
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_ForestMgmt to get the corresponding forest management types
  left_join(map_ForestMgmt %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  # Group by id_c and type to summarize total area per cover class per pixel
  group_by(id_c, type) %>%
  summarise(value = sum(AreaPerCover, na.rm = TRUE), .groups = 'drop') %>%
  # Pivot wider to create columns for each forest management type
  pivot_wider(names_from = type, values_from = value, values_fill = 0) 


#Mean Travel time
Xmat_Travel <- Travel.df %>% select(id_c, MeanTravelTime) %>% 
  mutate(log_MeanTravelTime = log(MeanTravelTime)) %>% 
  select(-MeanTravelTime)
#summary(Xmat_Travel)

#Population  
Xmat_pop <- population.df %>% select(id_c, TotalPop) %>% 
  mutate(log_pop = log(TotalPop)) %>% 
  select(-TotalPop)
#summary(Xmat_pop)  
  
#Altitude  
Xmat_alt <- altitude.df %>% select(id_c, MeanAltitude) %>% 
  mutate(log_altitude = log(MeanAltitude)) %>% 
  select(-MeanAltitude)

#Slope and data from GAEZ on zone and thermal 
Xmat_slope <- slope.df %>%
  select(id_c, MeanSlope, thermal, zone) %>%
  # Create binary columns for each thermal category
  pivot_wider(id_cols = c(id_c, MeanSlope), 
              names_from = thermal, 
              values_from = thermal, 
              names_prefix = "thermal_", 
              values_fn = list(thermal = ~ 1), 
              values_fill = list(thermal = 0)) %>%
  # Retain MeanSlope before the second pivot for zone
  left_join(
    slope.df %>%
      select(id_c, zone) %>%
      pivot_wider(id_cols = id_c, 
                  names_from = zone, 
                  values_from = zone, 
                  names_prefix = "zone_", 
                  values_fn = list(zone = ~ 1), 
                  values_fill = list(zone = 0)), 
    by = "id_c"
  ) %>%
  mutate(log_slope = log(MeanSlope)) %>%
  select(-MeanSlope)


Xmat <- Xmat_LU %>% 
  select(-forest_share) %>% 
  mutate(Intercept = 1) %>% 
  left_join(Xmat_Travel) %>% 
  left_join(Xmat_pop) %>% 
  left_join(Xmat_ForestManagement) %>% 
  left_join(Xmat_alt) %>% 
  left_join(Xmat_slope) %>% 
  select(-c(cropland, otherland, urban, pasture, forest)) %>% #, SRP_share)) %>% 
  #pivot back longer to fit the DownscalR package structure
  pivot_longer(-id_c, values_to = "value", names_to = "ks") %>% 
  rename(ns = id_c)

```

```{r changing land cover structure to create lu_levels, echo = FALSE, message = FALSE}

#land use level per cell per land cover class in baseline year

lu_levels <- LandCover.df %>%
  # Pivot to long format and convert AreaPerCover to numeric
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_Copernicus and filter out "ocean" and "water"
  left_join(map_Copernicus %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  filter(!LandCover %in% c("ocean", "water")) %>%
  # Group by id_c and LandCover, summarizing the total area per cover class
  group_by(id_c, LandCover) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha
  summarise(value = sum(AreaPerCover, na.rm = TRUE)/10, .groups = 'drop') %>%
  # Rename columns for clarity
  rename(ns = id_c, lu.from = LandCover) %>%
  # Add rows for "newforest" with value = 0 for each unique id_c
  bind_rows(data.frame(ns = unique(LandCover.df$id_c), lu.from = "newforest", value = 0))

```

```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}
#find where I have NAs in the Xmat matrix to remove those cells else the MNL cannot run
rm_na <- Xmat$ns[which(is.na(Xmat), arr.ind = T)[,1]]

country_df <- list(xmat = tibble(Xmat %>% 
                                   filter(!(ns %in% rm_na))), 
                   lu_levels = tibble(lu_levels %>% 
                                        filter(!(ns %in% rm_na))))
```

# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}

FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = pathway) %>%
  # In calculator, sheet 4_calc_land, 
  # table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  
  # Remove YearStart and reshape to long format
  select(-YearStart) %>%
  pivot_longer(-c(YearEnd, LandCoverInit), values_to = "value", names_to = "lu.to") %>%
  # Clean and adjust land cover names
  mutate(
    lu.from = tolower(LandCoverInit),
    lu.to = tolower(str_remove(lu.to, "To")),
    times = YearEnd
  ) %>%
  # Select necessary columns and filter out unchanged land covers
  select(times, lu.from, lu.to, value) %>%
  filter(lu.from != lu.to)
  

table(FABLE$lu.from, FABLE$lu.to)
```

```{r Harmonize targets and Land Cover Change baseline map}
country_LUC_ag <- LandCoverChange.df %>% 
  #creating the structure to have land cover change information per id
  select(id_c) %>% 
  mutate(temp = 1) %>% 
  full_join(map_HILDA%>%  mutate(code = as.character(code)) %>% mutate(temp = 1), relationship = "many-to-many") %>% 
  #merge with data on Land Cover Change per code
  left_join(LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA %>%  mutate(code = as.character(code)), by = c("name" = "code")),
  relationship = "many-to-many") %>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(AreaPerCode, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2018) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  group_by(lu.from, lu.to, Ts) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha
  summarise(value_HILDA = sum(value, na.rm = TRUE)/10, .groups = 'drop') %>% #/1000 to get same unit as FABLE targets
  rename(times = Ts) %>% 
  mutate(times = 2020)


Baseline_LU <- Xmat_LU %>% 
  select(id_c, cropland, forest, otherland, pasture, urban) %>% 
  pivot_longer(cols = c(cropland:urban), values_to = "value_Copernicus", names_to = "LandCover") %>% 
  group_by(LandCover) %>% 
  #this is in km2 so I divide by 10 to get in 1000 ha
  summarise(value_Copernicus = sum(value_Copernicus, na.rm = T)/10) %>% 
  pivot_wider(names_from = LandCover, values_from = value_Copernicus) %>% 
  mutate(year = 2019)


comp <- country_LUC_ag %>% 
  left_join(FABLE %>% 
              mutate(value = value/5) %>%
              rename(value_FABLE = value)) %>% 
  filter(lu.from!=lu.to)
```




# Estimate MNL model

```{r prep data, echo = FALSE, message=FALSE, include = FALSE}

results_MNL <- data.frame()
pred_coeff_long <- data.frame()
country_start_areas <- data.frame()
X_long <- data.frame()

Yraw <- country_luc %>% 
  filter(!(ns %in% rm_na))

for(cur_LU.from in c("urban", "cropland", "pasture", "forest", "otherland", "newforest")){
  
 Y <- dplyr::filter(Yraw,lu.from == cur_LU.from & Ts == 2018) %>% #changes 2018 to 2019
  tidyr::pivot_wider(names_from = lu.to) %>%
  tibble::column_to_rownames(var = "ns") %>%
  dplyr::select(-c(lu.from, Ts))
 
 X <- country_df$xmat %>% 
    tidyr::pivot_wider(names_from = "ks") %>%
    dplyr::arrange(match(ns,Y$ns)) %>%
    tibble::column_to_rownames(var = "ns") 
  
  baseline <- which(colnames(Y) == cur_LU.from)

  # Decreased niter = 100 and nburn = 50 for faster computation 
  temp_results_MNL <- mnlogit(as.matrix(X), as.matrix(Y), baseline, niter = 100, nburn = 50, A0 = 10^4)
  
  pred_coeff <- as.data.frame(apply(temp_results_MNL$postb[,-baseline,],c(1,2),mean))
  
  pred_coeff_long <- pred_coeff_long %>%
    rbind(
    pred_coeff %>% tibble::rownames_to_column(var="ks") %>%
  tidyr::pivot_longer(!ks, names_to = "lu.to", values_to="value") %>%
  dplyr::arrange(lu.to) %>%
  tibble::add_column(lu.from=cur_LU.from, .after="ks")
    )
  
  ## starting area should provide information on area of each (higher resolution) grid cell, i.e ns
country_start_areas <- country_start_areas %>%
  rbind(
    dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>% mutate(times = 2020)
  )

}
```

# Downscaling

```{r restriction data, echo = FALSE, message=FALSE}

## Protected areas

```



```{r, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}


X_long <-  X %>% tibble::rownames_to_column(var="ns") %>%
  tidyr::pivot_longer(!ns, names_to="ks", values_to = "value")

## Downscaling computation and results

results_DS <- downscale(targets = FABLE %>% filter(times > 2020), 
                        start.areas =  country_start_areas, 
                        xmat = X_long, 
                        betas = pred_coeff_long)

downscaled_LUC <- results_DS$out.res

```

# Land-use map

```{r, echo = FALSE, message=FALSE}
LUC_plot <- LUC_plot(results_DS, rasterized_layer)
plot_to_save <- LUC_plot$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)

```

# Land-use change map

```{r, echo = FALSE, message = FALSE}

results_DS_LUC <- results_DS

# Create 'gain' and 'loss' datasets
results_DS_LUC$out.res <- results_DS_LUC$out.res %>%
  filter(lu.from != lu.to) %>%
  # Group by `lu.to`, `ns`, and `times` for gains
  group_by(lu.to, ns, times) %>%
  summarise(gain = sum(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(Type = "gain", lu = lu.to) %>%
  
  # Bind the losses dataset by grouping by `lu.from`
  bind_rows(
    results_DS_LUC$out.res %>%
      filter(lu.from != lu.to) %>%
      group_by(lu.from, ns, times) %>%
      summarise(loss = -sum(value, na.rm = TRUE), .groups = "drop") %>%
      mutate(Type = "loss", lu = lu.from)
  ) %>%
  
  # Summarize the final dataset grouping by land-use (`lu`), time, and pixel (`ns`)
  group_by(lu, times, ns) %>%
  summarise(value = sum(gain, loss, na.rm = TRUE), .groups = "drop") %>%
  rename(lu.to = lu)

#Define plot aesthetics
color = "PRGn"
label = "Area gains and losses in ha per pixel"

#Create the transition plot using LUC_plot
Transition_plot <- LUC_plot(
  results_DS_LUC, rasterized_layer, 
  color = color,
  label = label
)

#Define the limit for color scaling
limit <- max(abs(Transition_plot[["plot.df"]]$value)) * c(-1, 1)

# Step 5: Create the plot with updated scale and theme adjustments
Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))

#Save the plot to a .tiff file
tiff(
  here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-change-map.tiff")),
  units = "in", height = 6, width = 10, res = 300
)
plot(Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))
)
dev.off()



```

```{r 2020 map, echo = FALSE, message = FALSE}

results_DS_2050 <- results_DS
results_DS_2050$out.res <- results_DS_2050$out.res %>% 
  dplyr::filter(times %in% c(2015, 2050))

# Create the land-use change plot for 2050
LUC_plot_2050 <- LUC_plot(results_DS_2050, rasterized_layer)
plot_to_save <- LUC_plot_2050$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "2050_Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)
```

