---
title: "Downscaling"
author: "Clara Douzal"
date: "2023-10-24"
output: pdf_document
params:
  variable1 : country
  variable2 : pathway 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
```


```{r temporary settings, include = FALSE}
#Please ensure that you have a folder named like the country variable you will input below in your "Output" folder. This is where your maps will be saved.

country = "IND"
pathway = "CurrentTrends"
# pathway = "NationalCommitments"
# pathway = "GlobalSustainability"

```

This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population <- geojson_read(here("Data", country, "Population2020.geojson"),  what = "sp")
population.df <- population@data %>% 
  select(-id) #%>% 
  #mutate(TotalPop = c(scale(TotalPop, scale = T)))

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"),  what = "sp")
ProtectedAreas.df <- ProtectedAreas@data %>% 
  select(-id)

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"),  what = "sp")
Travel.df <- Travel@data %>% 
  select(-id)

LandCover <- geojson_read(here("Data", country, "LandCoverCopernicus2019.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)

LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2018_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data %>% 
   select(-id)

ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data %>% 
  select(-id)

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")

map_ForestMgmt <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ForestManagement")
```

```{r create raster for maps, include = FALSE}
#raster to use for plots
ext <- extent(Travel)
res <- 0.5
raster_layer <- raster(ext, res = res)
rasterized_layer <- rasterize(Travel, raster_layer, field = "id")
```

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
country_luc <- LandCoverChange.df %>% 
  #creating the structure to have land cover change information per id
  select(id_c) %>% 
  mutate(temp = 1) %>% 
  full_join(map_HILDA%>%  mutate(code = as.character(code)) %>% mutate(temp = 1), relationship = "many-to-many") %>% 
  #merge with data on Land Cover Change per code
  left_join(LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA %>%  mutate(code = as.character(code)), by = c("name" = "code")),
  relationship = "many-to-many") %>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2018) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2018,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is posible in future even though not in the calibration
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2018,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is possible in future even though not in the calibration
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2018,
                   value = 0)) 

table(country_luc$lu.from, country_luc$lu.to)
  
```

```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}
#Right now only has land cover data but should contain all the explanatory variables to compute the priors
Xmat_LU <- LandCover.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover))  %>% 
  mutate(name = stringr::str_remove(name, "X")) %>% 
  #associate the right land cover names
  left_join(map_Copernicus %>%  mutate(code = as.character(code)), by = c("name" = "code")) %>% 
  group_by(id_c, LandCover) %>% 
  #compute the area per cover class in each cell
  summarise(value = sum(AreaPerCover, na.rm = T)) %>% 
  dplyr::select(id_c, LandCover, value) %>% 
  #pivot wider to add the intercept value
  pivot_wider(names_from = "LandCover", values_from = "value") %>% 
  #doesn't exist in copernicus and hilda need to add it
  #mutate(newforest = 1) %>% 
#To add for GBR to have a non singulary matrix
  select(-unknwon) %>% 
  #remove the water and ocean land cover
  select(-c(water, ocean)) %>% 
  #compute the share of each land use class within a pixel
  rowwise() %>%
  mutate(
    #making sure there is no NaN by setting the denominator to 1
          total = ifelse(sum(c_across(cropland:urban)) == 0, 
                            1,
                            sum(c_across(cropland:urban))),
         across(c(cropland, forest, otherland, pasture, urban),
                .fns = ~ .x / total, .names = "{.col}_share")) %>%
  ungroup() %>%
  select(-total)
# use summary to check if there are NAs in the Xmat
#summary(Xmat_LU)

#ForestManagement
Xmat_ForestManagement <- ForestManagement.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover))  %>% 
  mutate(name = stringr::str_remove(name, "X")) %>% 
  #associate the right land cover names
  left_join(map_ForestMgmt %>%  mutate(code = as.character(code)), by = c("name" = "code")) %>% 
  group_by(id_c, type) %>% 
  #compute the area per cover class in each cell
  summarise(value = sum(AreaPerCover, na.rm = T)) %>% 
  dplyr::select(id_c, type, value) %>% 
  #pivot wider to add the intercept value
  pivot_wider(names_from = "type", values_from = "value") %>%
  #compute the share of each forest management within a pixel, set to 0 if no forest in the pixel
  rowwise() %>%
  mutate(
    #making sure there is no NaN by setting the denominator to 1
    total = ifelse(sum(c_across(all_of(setdiff(names(.), "id_c")))) ==0,
                   1,
                   sum(c_across(all_of(setdiff(names(.), "id_c"))))),
    across(all_of(setdiff(names(.), "id_c")),
           .fns = ~ if_else(total == 0, 0, .x / total), .names = "{.col}_share")) %>%
  ungroup() %>%
  select(-total) %>%
  #add the share of forest in the pixel from LU.df
  left_join(Xmat_LU %>% select(id_c, forest_share)) %>% 
  #compute the share of forest under specific forest management practice in each pixel
  mutate(across(contains("share"), ~ .x * forest_share)) %>% 
  select(-forest_share) %>% 
  select(c(id_c, contains("share")))
#summary(Xmat_ForestManagement)

#Mean Travel time
Xmat_Travel <- Travel.df %>% select(id_c, MeanTravelTime) %>% 
  mutate(log_MeanTravelTime = log(MeanTravelTime)) %>% 
  select(-MeanTravelTime)
#summary(Xmat_Travel)

#Population  
Xmat_pop <- population.df %>% select(id_c, TotalPop) %>% 
  mutate(log_pop = log(TotalPop)) %>% 
  select(-TotalPop)
#summary(Xmat_pop)  
  

Xmat <- Xmat_LU %>% 
  select(-forest_share) %>% 
  mutate(Intercept = 1) %>% 
  left_join(Xmat_Travel) %>% 
  left_join(Xmat_pop) %>% 
  left_join(Xmat_ForestManagement) %>% 
  select(-c(cropland, otherland, urban, pasture, forest)) %>% #, SRP_share)) %>% 
  #pivot back longer to fit the DownscalR package structure
  pivot_longer(-id_c, values_to = "value", names_to = "ks") %>% 
  rename(ns = id_c)

```

```{r changing land cover structure to create lu_levels, echo = FALSE, message = FALSE}

#land use level per cell per land cover class in baseline year

lu_levels <- LandCover.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover))  %>% 
  mutate(name = stringr::str_remove(name, "X")) %>% 
  #associate the right land cover names
  left_join(map_Copernicus %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>% 
  filter(!(LandCover %in% c("ocean", "water"))) %>% 
  group_by(id_c, LandCover) %>% 
  #Compute the area per cover class per cell (i.e. this is not a share, should it be a share?)
  summarise(value = sum(AreaPerCover, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = LandCover) %>% 
  dplyr::select(ns, lu.from, value) %>% 
  rbind(data.frame(ns = unique(LandCover.df$id_c),
                   lu.from = "newforest",
                   value = 0)) 

```

```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}

rm_na <- Xmat$ns[which(is.na(Xmat), arr.ind = T)[,1]]

country_df <- list(xmat = tibble(Xmat %>% 
                                   filter(!(ns %in% rm_na))), 
                   lu_levels = tibble(lu_levels %>% 
                                        filter(!(ns %in% rm_na))))
```

# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}
FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"),
                            sheet = pathway) %>% #In calculator, sheet 4_calc_land, 
  #table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  dplyr::select(-YearStart) %>% 
  pivot_longer(-c(YearEnd, LandCoverInit), 
               values_to = "value", 
               names_to = "lu.to") %>% 
  mutate(lu.to = stringr::str_remove(lu.to, "To")) %>% 
  mutate(lu.from = tolower(LandCoverInit),
         lu.to = tolower(lu.to),
         times = YearEnd) %>% 
  dplyr::select(times, lu.from, lu.to, value) %>% 
  #do not keep land cover that remains the same
  filter(!(lu.from == lu.to))
  

table(FABLE$lu.from, FABLE$lu.to)
```

# Estimate MNL model

```{r prep data, echo = FALSE, message=FALSE, include = FALSE}

results_MNL <- data.frame()
pred_coeff_long <- data.frame()
country_start_areas <- data.frame()
X_long <- data.frame()

Yraw <- country_luc %>% 
  filter(!(ns %in% rm_na))

for(cur_LU.from in c("urban", "cropland", "pasture", "forest", "otherland", "newforest")){
  
 Y <- dplyr::filter(Yraw,lu.from == cur_LU.from & Ts == 2018) %>% #changes 2018 to 2019
  tidyr::pivot_wider(names_from = lu.to) %>%
  tibble::column_to_rownames(var = "ns") %>%
  dplyr::select(-c(lu.from, Ts))
 
 X <- country_df$xmat %>% 
    tidyr::pivot_wider(names_from = "ks") %>%
    dplyr::arrange(match(ns,Y$ns)) %>%
    tibble::column_to_rownames(var = "ns") 
  
  baseline <- which(colnames(Y) == cur_LU.from)

  # Decreased niter = 100 and nburn = 50 for faster computation 
  temp_results_MNL <- mnlogit(as.matrix(X), as.matrix(Y), baseline, niter = 100, nburn = 50, A0 = 10^4)
  
  pred_coeff <- as.data.frame(apply(temp_results_MNL$postb[,-baseline,],c(1,2),mean))
  
  pred_coeff_long <- pred_coeff_long %>%
    rbind(
    pred_coeff %>% tibble::rownames_to_column(var="ks") %>%
  tidyr::pivot_longer(!ks, names_to = "lu.to", values_to="value") %>%
  dplyr::arrange(lu.to) %>%
  tibble::add_column(lu.from=cur_LU.from, .after="ks")
    )
  
  ## starting area should provide information on area of each (higher resolution) grid cell, i.e ns
country_start_areas <- country_start_areas %>%
  rbind(
    dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>% mutate(times = 2020)
  )

}
```

# Downscaling

```{r restriction data, echo = FALSE, message=FALSE}

## Protected areas

```



```{r, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}


X_long <-  X %>% tibble::rownames_to_column(var="ns") %>%
  tidyr::pivot_longer(!ns, names_to="ks", values_to = "value")

## Downscaling computation and results

results_DS <- downscale(targets = FABLE %>% filter(times > 2020), 
                        start.areas =  country_start_areas, 
                        xmat = X_long, 
                        betas = pred_coeff_long)

downscaled_LUC <- results_DS$out.res

```

# Land-use map

```{r, echo = FALSE, message=FALSE}
LUC_plot <- LUC_plot(results_DS, rasterized_layer)
LUC_plot$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))

tiff(here("Output", country, paste0(format(Sys.Date(),format = "%y%m%d"), "_", pathway , "_", "Land-use-map.tiff")), units = "in", height = 6, width = 10, res = 300)
plot(LUC_plot$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm")))
dev.off()

```

# Land-use change map

```{r, echo = FALSE, message = FALSE}



results_DS_LUC <- results_DS
results_DS_LUC$out.res <- rbind(
  results_DS_LUC$out.res %>% 
  filter(lu.from != lu.to) %>% 
    arrange(lu.to, ns, times) %>% 
    group_by(lu.to, ns, times)%>% 
    summarise(value = sum(value, na.rm = T)) %>% 
    mutate(Type = "gain") %>% 
    rename(lu = lu.to),
  results_DS_LUC$out.res %>% 
  filter(lu.from != lu.to) %>% 
    arrange(lu.from, ns, times) %>% 
    group_by(lu.from, ns, times)%>% 
    summarise(value = -sum(value, na.rm = T)) %>% 
    mutate(Type = "loss") %>%
    rename(lu = lu.from)) %>% 
  group_by(lu, times, ns) %>% 
  #get net change from lu cat, for a cell for a year
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(lu.to = lu)
  

color = "PRGn"
label = "Area gains and losses in ha per pixel"
Transition_plot <- LUC_plot(results_DS_LUC, rasterized_layer, 
                            color = "PRGn",
                            label = "Area gains and losses in ha per pixel")

limit <- max(abs(Transition_plot[["plot.df"]]$value)) * c(-1, 1)
Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2()+
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))

tiff(here("Output", country, paste0(format(Sys.Date(),format = "%y%m%d"), "_", pathway , "_","Land-use-change-map.tiff")), units = "in", height = 6, width = 10, res = 300)
plot(Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2()+
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm")))
dev.off()
#palette = color, 
                                #name=label)#, 
                                #direction = 1)#, 
                                #limit = limit)


# plot_obj <- ggplot2::ggplot() +
#     ggplot2::geom_tile(data=Transition_plot[["plot.df"]], 
#                        aes(x=x, y=y, 
#                            # fill=factor(cut(value, breaks =  unique(quantile(value, probs = seq(0.,1,0.10))))),
#                            fill = factor(cut(value, 
#                                              breaks = fullseq(range(value),(range(value)[2]-range(value)[1])/8))),
#                            group=lu.to), 
#                        alpha=0.8) +
#     ggplot2::scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PRGn"),#palette = "PRGn",
#                                   #direction = 1,
#                                   #limit = limit#,
#                                   #name=label
#                                   name = NULL)+
#     ggplot2::coord_equal() +
#     ggthemes::theme_map() +
#     ggplot2::theme(legend.position="bottom") +
#     ggplot2::theme(legend.key.width=unit(2, "cm"))+
#     ggplot2::facet_grid(times~lu.to)
#   
# plot_obj

```

```{r 2020 map, echo = FALSE, message = FALSE}

#targets_2050 <- dplyr::filter(FABLE, times == 2050)

results_DS_2050 <- results_DS
results_DS_2050$out.res <- results_DS_2050$out.res %>% dplyr::filter(times %in% c(2015, 2050))

LUC_plot_2050 <- LUC_plot(results_DS_2050, rasterized_layer)
LUC_plot_2050$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm"))

tiff(here("Output", country, paste0(format(Sys.Date(),format = "%y%m%d"), "_", pathway , "_", "2050_Land-use-map.tiff")), units = "in", height = 6, width = 10, res = 300)
plot(LUC_plot_2050$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm")))
dev.off()


```

