---
title: "Downscaling"
author: "Clara Douzal"
date: "2023-10-24"
output: pdf_document
header-includes:
  - \DeclareUnicodeCharacter{2061}{}
  - \DeclareUnicodeCharacter{2212}{}
params:
  country : "IND"
  pathway : "GlobalSustainability"
  mnl_niter : 100
  mnl_nburn : 50
  seed : 1212
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 150)

```



```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
library(stringr)
#library(kableExtra)
library(knitr)
library(readr)

if (requireNamespace("conflicted", quietly = TRUE)) {
  conflicted::conflict_prefer("filter", "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("select", "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("lag", "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("rename", "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("left_join", "dplyr", quiet = TRUE)
}
```

```{r setting paths, include = FALSE}
set.seed(params$seed)
country <- params$country
pathway <- params$pathway
niter <- params$mnl_niter
nburn <- params$mnl_nburn

dir.create(here("Output", country), showWarnings = FALSE, recursive = TRUE)
stamp <- format(Sys.Date(), "%y%m%d")
tag <- paste(stamp, country, pathway, sep = "_")

```


This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE, cache = TRUE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population <- geojson_read(here("Data", country, "Population2020.geojson"),  what = "sp")
population.df <- population@data %>% 
  select(-id) #%>% 
  #mutate(TotalPop = c(scale(TotalPop, scale = T)))

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"),  what = "sp")
ProtectedAreas.df <- ProtectedAreas@data %>% 
  select(-id)

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"),  what = "sp")
Travel.df <- Travel@data %>% 
  select(-id)

#this is in km2 
LandCover <- geojson_read(here("Data", country, "LandCoverHILDA2015.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)

#this is in km2
LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2015_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data %>% 
   select(-id)

OldLandCoverChange <- geojson_read(here("Data", country, "OLD GRID","LandCoverChangeHILDA2015_2019.geojson"),  what = "sp")
OldLandCoverChange.df <- OldLandCoverChange@data %>% 
   select(-id)


ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data %>% 
  select(-id)

altitude <- geojson_read(here("Data", country, "Altitude.geojson"),  what = "sp")
altitude.df <- altitude@data %>% 
  select(-id)

slope <- geojson_read(here("Data", country, "Slope.geojson"),  what = "sp")
slope.df <- slope@data %>% 
  select(-id)

gaez <- geojson_read(here("Data", country, "GAEZCropDistribution2015.geojson"),  what = "sp")
gaez.df <- gaez@data  %>% 
  select(-id) %>% 
  select(-total)

livestock <- geojson_read(here("Data", country, "GLW4WorldGriddedLivestock2020.geojson"),  what = "sp")
livestock.df <- livestock@data  %>% 
  select(-id) %>% 
  select(-alllivestock)

#Loading the 50km grid with
grid <- read_csv(here("Data", "global", "grid50_equal_area.csv")) %>% 
  filter(iso3 == country)

zones <- grid %>% 
  select(id_c, thermal, zone, BIOME_NAME)

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_HILDA_LUC <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA_change")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")

map_ForestMgmt <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ForestManagement")
```

Below is the table of input data used in the prior module of the downscaling process.

```{r echo=FALSE, message=FALSE}
library(knitr)

# Create the data frame for the table
table_data <- data.frame(
  `Variable in econometric model` = c("Dependent", "Land use change", "", 
                                      "Explanatory variables", "Land use", 
                                      "Forest Management", "Mean Travel time to major cities", "Population"),
  Year = c("", "2015-2019", "", "", "2015", "2015", "2000", "2020"),
  Unit = c("", "% of lu.from in pixel being converted", "", "", 
           "hectares", "hectares", "minutes", "Count"),
  Source = c("", "HILDA", "", "", "COPERNICUS", "Lesiv et al.", 
             "Nelson, A (2008)", "WorldPop")
)

# Use kable without booktabs
kable(
  table_data, 
  format = "latex", 
  col.names = c("Variable in econometric model", "Year", "Unit", "Source")
)
```


## Checking the land cover change input data

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix. With "missing_from" you can see the number of land transitions thatare missing because of a missing initial land cover. With "missing_to" you can see the number of land transitions that are missing because of a missing final land cover.


```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
#value is the percentage of the cell that is going from land cover to another
country_luc <- LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA_LUC %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2015) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2015,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) 

# table(country_luc$lu.from, country_luc$lu.to)
  
expected <- c("cropland","forest","newforest","otherland","pasture","urban")
present_from <- sort(unique(country_luc$lu.from))
present_to   <- sort(unique(country_luc$lu.to))
data.frame(kind = c("missing_from","missing_to"),
           value = I(list(setdiff(expected, present_from),
                         setdiff(expected, present_to)))) %>% knitr::kable()
```

If any land transitions are missing you can check the table below.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r , echo = FALSE, message = FALSE}
table(country_luc$lu.from, country_luc$lu.to) %>% knitr::kable()
```

# Harmonizing start map with FABLE Calculator baseline map

We need to harmonize the high-resolution baseline map used in the downscalR function with the aggregated land-use data from the FABLE Calculator. This ensures that the spatially explicit baseline map aligns with the aggregated land-use data in FABLE. Without this alignment, there could be discrepancies in the starting levels of each land-use class, making the future land-use transitions projected by the FABLE Calculator unrealistic when applied to the unadjusted high-resolution map.

```{r changing land cover structure to create start_map harmonized with FABLE aggregated land cover, echo = FALSE, message = FALSE}

#land use level per cell per land cover class in baseline year
start_map <- LandCover.df %>%
  # Pivot to long format and convert AreaPerCover to numeric
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_Copernicus and filter out "ocean" and "water"
  left_join(map_HILDA %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  filter(!LandCover %in% c("ocean", "water")) %>%
  # Group by id_c and LandCover, summarizing the total area per cover class
  group_by(id_c, LandCover) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha - 1 km² = 0.1 thousand ha
  summarise(value = sum(AreaPerCover, na.rm = TRUE)/10, .groups = 'drop') %>%
  # Rename columns for clarity
  rename(ns = id_c, lu.from = LandCover) %>%
  # Add rows for "newforest" with value = 0 for each unique id_c
  bind_rows(data.frame(ns = unique(LandCover.df$id_c), lu.from = "newforest", value = 0)) %>% 
  mutate(lu.from = ifelse(is.na(lu.from), "not relevant", lu.from))
  
start_map_harm <- start_map %>% 
  group_by(lu.from) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(lu = lu.from) %>% 
  mutate(lu = ifelse(is.na(lu), "not relevant", lu))

#FABLE Calculator aggregated land use level in baseline year
LC_targets <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = "Baseline") %>% 
  rename(lu = LandCover)

keep <- "from" # Specify which areas to keep during transition

# Calculate land-use change targets based on the start and target data
LUC_targets <- LU_to_LUC(start_map_harm, 
                              LC_targets, 
                              keep_areas = keep)

# Create a check data frame to validate the results
check_df <- rbind(NULL,
                  left_join(
                    start_map_harm %>% group_by(lu) %>% 
                      rename(lu.from = "lu") %>% 
                      reframe(value_start = sum(value)), 
                    LC_targets %>% group_by(lu) %>% 
                      rename(lu.to = "lu") %>% 
                      reframe(value_target = sum(value)), 
                    by = c("lu.from" = "lu.to")) %>%
                    left_join(LUC_targets %>% 
                                group_by(lu.to) %>% 
                                rename(value_new = "value") %>% 
                                reframe(value_new = sum(value_new)), 
                              by = c("lu.from" = "lu.to")) %>% 
                    select(lu.from, value_start, value_target, value_new) %>%
                    mutate(new_target_diff = value_new - value_target, 
                           value_rebalance = sum(value_start) * (value_target / sum(value_target)), 
                           rebalance_new_diff = value_rebalance - value_new))

### Calculate implied transition shares from LUC targets
CLC_FABLE_shares <- LUC_targets %>% 
  group_by(lu.from) %>% 
  mutate(value = value / sum(value)) %>% # Normalize values to get shares
  pivot_wider(id_cols = c(lu.from), names_from = "lu.to", values_from = "value") %>% 
  replace(is.na(.), 0) %>% # Replace NA with 0 for clarity
  pivot_longer(cols = !c(lu.from), names_to = "lu.to", values_to = "proj_share")

# Apply the transition shares to the high-resolution start map for downscaling
start_map_reproj <- left_join(start_map, CLC_FABLE_shares, relationship = "many-to-many") %>% 
  mutate(value = ifelse(!is.na(lu.to), value * proj_share, value), # Adjust values based on shares
         lu.to = ifelse(is.na(lu.to), lu.from, lu.to)) %>% # Update land use type
  group_by(ns, lu.to) %>% 
  summarize(value = sum(value)) %>% # Sum values for each unique combination
  ungroup() %>% 
  drop_na() %>% # Remove rows with NA values
  rename(lu.from = "lu.to") # Rename for consistency


```

```{r create a stable ID map, include = FALSE}

# === build a master ID map (character id -> integer code) ===
# start_map already has ns (from id_c)
ns_map <- start_map %>%
  dplyr::distinct(ns) %>%
  dplyr::mutate(ns = as.character(ns)) %>%
  dplyr::arrange(ns) %>%
  dplyr::mutate(ns_int = dplyr::row_number(),
                id_c   = ns)  # convenience column for joining to sp@data$id_c

# sanity: no duplicates
stopifnot(!anyDuplicated(ns_map$ns))

# === rasterize *now* (doesn't need results yet) ===
# make sure Travel has id_c and same CRS
Travel@data <- Travel@data %>%
  dplyr::mutate(id_c = as.character(id_c)) %>%
  dplyr::left_join(ns_map, by = "id_c")

# fail fast if some polygons aren't in the map
if (anyNA(Travel@data$ns_int)) {
  missing <- unique(Travel@data$id_c[is.na(Travel@data$ns_int)])
  stop(sprintf("Travel polygons not found in ns_map (n=%d). Example: %s",
               length(missing), paste(head(missing, 5), collapse=", ")))
}

raster_layer <- raster::raster(raster::extent(Travel), res = 0.5, crs = raster::crs(Travel))
rasterized_layer <- raster::rasterize(Travel, raster_layer, field = "ns_int")

```
# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}

FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = pathway) %>%
  # In calculator, sheet 4_calc_land, 
  # table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  
  # Remove YearStart and reshape to long format
  select(-YearStart) %>%
  pivot_longer(-c(YearEnd, LandCoverInit), values_to = "value", names_to = "lu.to") %>%
  # Clean and adjust land cover names
  mutate(
    lu.from = tolower(LandCoverInit),
    lu.to = tolower(str_remove(lu.to, "To")),
    times = YearEnd
  ) %>%
  # Select necessary columns and filter out unchanged land covers
  select(times, lu.from, lu.to, value) %>%
  filter(lu.from != lu.to)
  
table(FABLE$lu.from, FABLE$lu.to) %>% knitr::kable()



```


```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}

Xmat_LU <- start_map_reproj %>% 
  pivot_wider(names_from = "lu.from", values_from = "value", values_fill = 0) %>% 
  rename(id_c = ns)

#ForestManagement
Xmat_ForestManagement <- ForestManagement.df %>%
  # Pivot longer to convert cover columns to rows
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_ForestMgmt to get the corresponding forest management types
  left_join(map_ForestMgmt %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  # Group by id_c and type to summarize total area per cover class per pixel
  group_by(id_c, type) %>%
  summarise(value = sum(AreaPerCover, na.rm = TRUE), .groups = 'drop') %>%
  # Pivot wider to create columns for each forest management type
  pivot_wider(names_from = type, values_from = value, values_fill = 0) 


#Mean Travel time
Xmat_Travel <- Travel.df %>% select(id_c, MeanTravelTime) %>% 
  transmute(id_c, log_MeanTravelTime = log1p(MeanTravelTime)) 


#Population  
Xmat_pop <- population.df %>% select(id_c, TotalPop) %>% 
  transmute(id_c, log_pop = log1p(TotalPop)) 
 
  
#Altitude  
Xmat_alt <- altitude.df %>% select(id_c, MeanAltitude) %>% 
  transmute(id_c, log_altitude = log1p(MeanAltitude))

Xmat_slope <- slope.df %>%
  dplyr::transmute(
    id_c     = as.character(id_c),
    log_slope = log1p(MeanSlope))

Xmat_zones <- zones %>% 
  dplyr::transmute(
    id_c     = as.character(id_c),
    thermal  = forcats::fct_explicit_na(as.factor(thermal), na_level = "UNK"),
    zone     = forcats::fct_explicit_na(as.factor(zone),    na_level = "UNK")
  ) %>%
  { 
    MM <- model.matrix(~ thermal + zone, data = .)   # has (Intercept) + k-1 dummies
    MM <- MM[, setdiff(colnames(MM), "(Intercept)"), drop = FALSE]  # drop intercept
    cbind(id_c = .$id_c, as.data.frame(MM))
  } %>%
  tibble::as_tibble() %>%
  # full coverage of pixels
  #dplyr::right_join(tibble::tibble(id_c = all_ns), by = "id_c") %>%
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("thermal"), ~ tidyr::replace_na(.x, 0)),
    dplyr::across(dplyr::starts_with("zone"),    ~ tidyr::replace_na(.x, 0))
  )

Xmat_gaez <- gaez.df %>% 
  mutate(id_c     = as.character(id_c)) %>%
  #drop columns that are all zeros
  select(
    any_of("id_c"),                               # keep id_c if present
    where(~ !is.numeric(.x) || any(abs(.x) > 1e-9, na.rm = TRUE))  # keep non-numeric OR has any non-zero
  )

# see which columns were dropped
# dropped_cols <- setdiff(names(gaez.df), names(Xmat_gaez))
# print(dropped_cols)

Xmat_livestock <- livestock.df %>% 
  mutate(id_c     = as.character(id_c)) %>% 
  #drop columns that are all zeros
  select(
    any_of("id_c"),                               # keep id_c if present
    where(~ !is.numeric(.x) || any(abs(.x) > 1e-9, na.rm = TRUE))  # keep non-numeric OR has any non-zero
  )
# see which columns were dropped
# dropped_cols <- setdiff(names(livestock.df), names(Xmat_livestock))
# print(dropped_cols)

Xmat <- Xmat_LU %>%
  mutate(Intercept = 1) %>%
  left_join(Xmat_Travel,           by = "id_c") %>%
  left_join(Xmat_pop,              by = "id_c") %>%
  left_join(Xmat_ForestManagement, by = "id_c") %>%
  left_join(Xmat_alt,              by = "id_c") %>%
  left_join(Xmat_slope,            by = "id_c") %>%
  left_join(Xmat_zones,            by = "id_c") %>% 
  left_join(Xmat_gaez,             by = "id_c") %>% 
  left_join(Xmat_livestock,        by = "id_c") %>% 
  pivot_longer(-id_c, names_to = "ks", values_to = "value") %>%
  rename(ns = id_c)


```

We cannot keep the cells that have NAs for some variables that will be used as priors.

```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}
#find where I have NAs in the Xmat matrix to remove those cells else the MNL cannot run
# rm_na <- Xmat$ns[which(is.na(Xmat), arr.ind = T)[,1]]
# 
# country_df <- list(xmat = tibble(Xmat %>% 
#                                    filter(!(ns %in% rm_na))), 
#                    lu_levels = tibble(start_map_reproj %>% 
#                                         filter(!(ns %in% rm_na))))

# After you build Xmat (long: ns, ks, value)
na_by_var <- Xmat %>%
  group_by(ks) %>%
  summarise(n_na = sum(is.na(value)), .groups="drop") %>%
  arrange(desc(n_na))

na_by_ns <- Xmat %>%
  group_by(ns) %>%
  summarise(n_na = sum(is.na(value)), .groups="drop") %>%
  arrange(desc(n_na))

head(na_by_var, 40) %>% knitr::kable()# which variables have NAs?
head(na_by_ns, 20) %>% knitr::kable()   # which cells have most NAs?


good_ns <- Xmat %>% 
  group_by(ns) %>% 
  summarise(ok = all(!is.na(value))) %>% 
  filter(ok) %>%
  pull(ns)

# Which cells were removed?
all_ns <- sort(unique(Xmat$ns))
removed_ns <- setdiff(all_ns, good_ns)

removal_msg <- if (length(removed_ns) == 0) {
  sprintf("**No grid cells were removed**: all %d cells have complete covariates.", length(all_ns))
} else {
  ids <- paste(head(removed_ns, 20), collapse = ", ")
  extra <- if (length(removed_ns) > 20) " ..." else ""
  sprintf("**%d grid cell(s) were removed** due to missing covariates. Removed cell IDs (showing up to 20): %s%s",
          length(removed_ns), ids, extra)
}

country_df <- list(
  xmat = Xmat %>% filter(ns %in% good_ns),
  lu_levels = start_map_reproj %>% filter(ns %in% good_ns)
)


saveRDS(country_df, here("Output", country, paste0(tag, "_countrydf.rds")))
```
`r removal_msg`

# Estimate MNL model


```{r prep data, echo = FALSE, message=FALSE, include = FALSE, cache = TRUE}
Yraw <- country_luc %>% 
  filter(ns %in% good_ns) #%>% 
  # filter(!(lu.to %in% c("not relevant", "water", "forest"))) %>% 
  # filter(!(lu.from %in% c("newforest", "urban", "not relevant")))

# 1) ONE global wide X for modeling (all cells, all ks)
X_wide <- country_df$xmat %>%
  tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
  dplyr::arrange(ns) %>%
  tibble::column_to_rownames(var = "ns")

# Optional safety check: no NAs
stopifnot(!any(is.na(X_wide)))

results_MNL <- data.frame()
pred_coeff_long <- data.frame()
country_start_areas <- data.frame()

for (cur_LU.from in c("urban","cropland","pasture","forest","otherland","newforest")) {

  Y <- dplyr::filter(Yraw, lu.from == cur_LU.from & Ts == 2015) %>%
    tidyr::pivot_wider(names_from = lu.to, values_from = value, values_fill = 0) %>%
    tibble::column_to_rownames(var = "ns") %>%
    dplyr::select(-lu.from, -Ts)

  # 2) Subset *rows only* from the global X_wide to match Y's ns
  X <- X_wide[rownames(Y), , drop = FALSE]

  baseline <- which(colnames(Y) == cur_LU.from)

  temp_results_MNL <- mnlogit(as.matrix(X),
                              as.matrix(Y),
                              baseline,
                              niter = niter, nburn = nburn, A0 = 1e4)

  pred_coeff <- as.data.frame(apply(temp_results_MNL$postb[,-baseline,], c(1,2), mean))

  pred_coeff_long <- pred_coeff_long %>%
    dplyr::bind_rows(
      pred_coeff %>%
        tibble::rownames_to_column(var = "ks") %>%
        tidyr::pivot_longer(!ks, names_to = "lu.to", values_to = "value") %>%
        dplyr::arrange(lu.to) %>%
        tibble::add_column(lu.from = cur_LU.from, .after = "ks")
    )

  country_start_areas <- country_start_areas %>%
    dplyr::bind_rows(
      dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>%
        dplyr::mutate(times = 2020)
    )
}

# 3) TRUE fix: X_long must come from the full, long xmat (NOT from X)
X_long <- country_df$xmat %>%
  dplyr::arrange(ns, ks)

# Optional safety: ensure ks in betas and X_long line up
stopifnot(all(unique(pred_coeff_long$ks) %in% unique(X_long$ks)))
stopifnot(all(unique(X_long$ks) %in% unique(pred_coeff_long$ks)))

# Downscaling
results_DS <- downscale(
  targets     = FABLE %>% dplyr::filter(times > 2020),
  start.areas = country_start_areas,
  xmat        = X_long,
  betas       = pred_coeff_long
)

# convert results ns -> ns_int using the same map
results_DS_num <- results_DS
results_DS_num$out.res <- results_DS$out.res %>%
  dplyr::mutate(ns = as.character(ns)) %>%
  dplyr::left_join(ns_map %>% dplyr::select(ns, ns_int), by = "ns") %>%
  dplyr::mutate(ns = ns_int) %>%
  dplyr::select(-ns_int)

stopifnot(!anyNA(results_DS_num$out.res$ns))

downscaled_LUC <- results_DS_num$out.res

```


# Downscaling - diagnostics & quality indicators

```{r restriction data, echo = FALSE, message=FALSE}

### code to use when ready to add constraints ####

## Protected areas

# Build a long mask of allowed = 0/1 per (ns, lu.to)
# Example: forbid cropland expansion inside ProtectedAreas
# pa_cells <- ProtectedAreas.df$id_c
# 
# restr <- expand.grid(ns = unique(start_map$ns),
#                      lu.to = c("cropland","forest","newforest","otherland","pasture","urban"),
#                      stringsAsFactors = FALSE) %>%
#   mutate(value = 1L) %>%
#   mutate(value = ifelse(ns %in% pa_cells & lu.to %in% c("cropland","urban","pasture"), 0L, value))
# 
# # Option A: If supported, pass to downscale()
# results_DS <- downscale(
#   targets = FABLE %>% filter(times > 2020),
#   start.areas = country_start_areas,
#   xmat = X_long,
#   betas = pred_coeff_long,
#   restriction = restr   # <-- if the API supports it
# )
# 
# saveRDS(results_DS, here("Output", country, paste0(tag, "_downscale.rds")))
# saveRDS(pred_coeff_long, here("Output", country, paste0(tag, "_mnl_betas.rds")))

```

## Aggregate consistency & diagonal zeros

This table compares the downscaled allocation with the national transition targets from the FABLE Calculator for each origin–destination land-use pair and year. For every (lu.from -> lu.to, time) it shows the allocated area, the target area, and a percentage error computed as 100 * (allocated − target) divided by the target (with a very small number added to avoid division by zero). 
How to interpret the results? Values should be very close to 0%. Systematically positive errors mean the optimizer allocates more than requested; negative errors mean less. Large discrepancies usually signal a mismatch in years, missing transitions in the targets, or binding restrictions that prevent exact matching.

```{r diagnostics & quality indicators - Aggregate consistency & diagonal zeros, echo=FALSE}

# Aggregate allocation vs targets
alloc_agg <- downscaled_LUC %>%
  group_by(times, lu.from, lu.to) %>%
  summarize(alloc = sum(value, na.rm=TRUE), .groups="drop")

targ_agg <- FABLE %>% 
  filter(times > 2020) %>% 
  mutate(times = as.character(times)) %>% 
  group_by(times, lu.from, lu.to) %>%
  summarize(target = sum(value, na.rm=TRUE), .groups="drop")  
  

consistency <- full_join(alloc_agg, targ_agg, by = c("times","lu.from","lu.to")) %>%
  mutate(across(c(alloc,target), ~ replace_na(.x, 0)),
         pct_err = 100 * (alloc - target) / pmax(abs(target), 1e-9)) %>% 
  filter(`lu.from` != `lu.to`)

knitr::kable(consistency %>% filter(times >= min(times))) 


```



## Prior vs allocation alignment

This diagnostic compares the priors implied by the multinomial model (what the econometrics expects) with the actual allocation from the optimization (what downscalR did).

Top-20 percent capture: rank cells by the prior for a given transition and report the fraction of allocated area that falls into the top 20 percent prior cells. 
How to interpret the results? A random allocation would be about 0.20. Higher is better (for example, 0.50 is good, 0.70 is very strong). Values near 0.20 can mean the priors are flat or that strong constraints overrode them.

Divergence measures: summary numbers such as Kullback–Leibler or Jensen–Shannon divergence quantify how far the allocation distribution is from the prior distribution across cells. 
How to interpret the results? Lower is better; 0 means perfect agreement.

Rank correlation: a Spearman correlation between prior scores and allocated intensities. Interpretation: above 0.6 is strong monotone agreement; 0.3 to 0.6 is moderate; near 0 is weak.

Note: when priors are intentionally flat (because of limited data), low capture or high divergence does not necessarily indicate an error.

```{r compute true prior probabilities for QA (KL & Top-20 capture), echo = FALSE}


# ---- Prep (run once after you have country_df, pred_coeff_long, results_DS) ----
# Wide design matrix aligned to feature names (ks)
X_wide <- country_df$xmat %>%
  tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
  dplyr::arrange(ns)
X_mat   <- as.matrix(X_wide %>% dplyr::select(-ns))
features <- colnames(X_mat)

# Destination classes (columns) — from your data
classes <- sort(unique(country_luc$lu.to))

# Numerically safe KL
kl_div <- function(q, p, eps = 1e-12) {
  q <- pmax(q, eps); p <- pmax(p, eps)
  sum(q * log(q / p))
}

# Optional: Jensen-Shannon divergence (symmetric, bounded)
js_div <- function(p, q, eps = 1e-12) {
  p <- pmax(p, eps); q <- pmax(q, eps)
  m <- 0.5*(p+q)
  0.5*kl_div(p, m, eps) + 0.5*kl_div(q, m, eps)
}

# Reconstruct prior probabilities for an origin class (lu.from)
prior_probs_for <- function(lu_from) {
  # Make a (features x classes) coefficient matrix with zeros by default.
  B <- matrix(0, nrow = length(features), ncol = length(classes),
              dimnames = list(features, classes))
  # Fill with estimated betas (long -> scatter into B)
  cf <- pred_coeff_long %>% dplyr::filter(lu.from == !!lu_from)
  ii <- match(cf$ks, features)
  jj <- match(cf$lu.to, classes)
  keep <- which(!is.na(ii) & !is.na(jj))
  if (length(keep)) B[cbind(ii[keep], jj[keep])] <- cf$value[keep]
  # Baseline column (lu_from) remains zeros  - > correct for multinomial logit.
  LP    <- X_mat %*% B
  maxr  <- apply(LP, 1, max)            # stabilize softmax
  expLP <- exp(LP - maxr)
  probs <- expLP / rowSums(expLP)
  list(ns = X_wide$ns, classes = classes, probs = probs)
}

# Compute KL & Top-k capture for one transition and one year
compute_transition_qc <- function(lu_from, lu_to, year) {
  pp  <- prior_probs_for(lu_from)
  j   <- match(lu_to, pp$classes)
  stopifnot(!is.na(j))
  p   <- as.numeric(pp$probs[, j])       # prior per cell for (lu_from -> lu_to)
  # Posterior allocation q from results (that year only)
  q_tbl <- results_DS$out.res %>%
    dplyr::filter(times == !!year, lu.from == !!lu_from, lu.to == !!lu_to) %>%
    dplyr::group_by(ns) %>% dplyr::summarise(area = sum(value), .groups="drop")
  q <- rep(0, length(pp$ns))
  q[match(q_tbl$ns, pp$ns)] <- q_tbl$area
  p <- p / pmax(sum(p), 1e-12)
  q <- q / pmax(sum(q), 1e-12)
  # Metrics
  Dkl <- kl_div(q, p)
  Jsd <- js_div(p, q)
  rho <- suppressWarnings(stats::cor(p, q, method = "spearman", use = "complete.obs"))
  # Top-20/40
  rk <- rank(-p, ties.method = "average")
  n  <- length(rk)
  top_capture <- function(frac) sum(q[rk <= frac*n])
  data.frame(
    year = year,
    lu.from = lu_from,
    lu.to = lu_to,
    KL_q_p = Dkl,
    JSD = Jsd,
    Spearman = rho,
    Top20 = top_capture(0.20)
   
  )
}

yrs <- sort(unique(results_DS$out.res$times))
base_year  <- min(yrs)
target_yr  <- if (2050 %in% yrs) 2050 else max(yrs)

#full QC table for all transitions in a year
all_pairs <- expand.grid(
  lu.from = c("cropland","forest","newforest","otherland","pasture","urban"),
  lu.to   = c("cropland","forest","newforest","otherland","pasture","urban"),
  stringsAsFactors = FALSE
) %>% dplyr::filter(lu.from != lu.to)

qc_table <- purrr::map_dfr(1:nrow(all_pairs), function(i) {
  with(all_pairs[i,], compute_transition_qc(lu.from, lu.to, target_yr))
})

knitr::kable(qc_table %>% dplyr::arrange(KL_q_p))

# How to read the numbers (cheat-sheet)
# 
# Top20 ~ 0.20 -> Allocation is basically random with respect to the prior.
# Top20 >= 0.50 -> Strong alignment to prior hotspots.
# Top20 >= 0.70 -> Very strong; most mass sits where priors said it should.
# 
# KL (q||p): 0 means perfect match. There isn’t a universal “good” threshold because it depends on sparsity and number of cells, but:
# 
# 0.00–0.10: extremely close to priors
# 
# 0.10–0.50: modest deviations
# 
# >0.50: sizable deviation; check constraints or priors (may be flat/weak)
# 
# JSD (optional): symmetric and bounded in 
# [0,ln⁡2]
# [0,ln2] ~ 
# [0,0.693]
# [0,0.693]. Interprets like KL (lower is better), but easier to compare across cases.
# 
# Spearman: rank agreement between prior and allocation across cells.
# ~0 means weak monotone relation; >0.4 moderate; >0.6 strong.
# 
# Tip: If priors are very flat (little contrast), KL can be small and Top20 near 0.2 even if the map “looks okay”. That’s a prior issue—consider strengthening drivers or adding constraints.

```



# Land-use map


```{r, echo = FALSE, message=FALSE, fig.align = "center", fig.width=8, fig.height=8, out.width='100%'}
# plot with consistent IDs on both sides
LUC_plot_out <- LUC_plot(results_DS_num, rasterized_layer)
plot_to_save <- LUC_plot_out$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)

```

# Land-use change map

```{r, echo = FALSE, message = FALSE, fig.width=8, fig.height=8, out.width='100%'}

results_DS_LUC <- results_DS_num

# Create 'gain' and 'loss' datasets
results_DS_LUC$out.res <- results_DS_LUC$out.res %>%
  filter(lu.from != lu.to) %>%
  # Group by `lu.to`, `ns`, and `times` for gains
  group_by(lu.to, ns, times) %>%
  summarise(gain = sum(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(Type = "gain", lu = lu.to) %>%
  
  # Bind the losses dataset by grouping by `lu.from`
  bind_rows(
    results_DS_LUC$out.res %>%
      filter(lu.from != lu.to) %>%
      group_by(lu.from, ns, times) %>%
      summarise(loss = -sum(value, na.rm = TRUE), .groups = "drop") %>%
      mutate(Type = "loss", lu = lu.from)
  ) %>%
  
  # Summarize the final dataset grouping by land-use (`lu`), time, and pixel (`ns`)
  group_by(lu, times, ns) %>%
  summarise(value = sum(gain, loss, na.rm = TRUE), .groups = "drop") %>%
  rename(lu.to = lu)

#Define plot aesthetics
color = "PRGn"
label = "Area gains and losses in 1000 ha per pixel"

#Create the transition plot using LUC_plot
Transition_plot <- LUC_plot(
  results_DS_LUC, rasterized_layer, 
  color = color,
  label = label
)

#Define the limit for color scaling
limit <- max(abs(Transition_plot[["plot.df"]]$value)) * c(-1, 1)

# Step 5: Create the plot with updated scale and theme adjustments
Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))

#Save the plot to a .tiff file
tiff(
  here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-change-map.tiff")),
  units = "in", height = 6, width = 10, res = 300
)
plot(Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))
)
dev.off()



```

```{r 2020 map, echo = FALSE, message = FALSE}

results_DS_2050 <- results_DS_num
results_DS_2050$out.res <- results_DS_2050$out.res %>% 
  dplyr::filter(times %in% c(2020, 2050))

# Create the land-use change plot for 2050
LUC_plot_2050 <- LUC_plot(results_DS_2050, rasterized_layer)
plot_to_save <- LUC_plot_2050$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "2050_Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)
```

## Priors check

### Standardized coefficients
```{r}
# Prep (once): wide X and feature SDs
X_wide <- country_df$xmat |>
  tidyr::pivot_wider(names_from = ks, values_from = value, values_fill = 0) |>
  dplyr::arrange(ns)
X_mat <- as.matrix(X_wide |> dplyr::select(-ns))
sd_x  <- apply(X_mat, 2, stats::sd)

classes <- sort(unique(country_luc$lu.to))

std_importance <- function(lu_from) {
  cf <- pred_coeff_long |> dplyr::filter(lu.from == !!lu_from)
  cf <- cf |> dplyr::mutate(sd = sd_x[match(ks, names(sd_x))],
                            beta_std = value * sd)
  # aggregate importance per feature across destination classes
  agg <- cf |>
    dplyr::group_by(ks) |>
    dplyr::summarise(
      max_abs_beta_std = max(abs(beta_std), na.rm = TRUE),
      l2_beta_std = sqrt(sum(beta_std^2, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(l2_beta_std))
  agg
}

# Example:
std_importance("otherland") |> head(15)  # top 15 features

```

### Average Marginal Effects (AMEs) on probabilities

```{r}
# Build coefficient matrix for an origin class (baseline column = origin; zeros)
coef_matrix_for <- function(lu_from) {
  B <- matrix(0, nrow = ncol(X_mat), ncol = length(classes),
              dimnames = list(colnames(X_mat), classes))
  cf <- pred_coeff_long |> dplyr::filter(lu.from == !!lu_from)
  ii <- match(cf$ks, rownames(B)); jj <- match(cf$lu.to, colnames(B))
  keep <- which(!is.na(ii) & !is.na(jj))
  if (length(keep)) B[cbind(ii[keep], jj[keep])] <- cf$value[keep]
  B
}

softmax <- function(M) {
  # M: N x J linear predictors
  M <- sweep(M, 1, apply(M, 1, max), FUN = "-")
  E <- exp(M)
  E / rowSums(E)
}

average_marginal_effects <- function(lu_from) {
  B  <- coef_matrix_for(lu_from)
  LP <- X_mat %*% B            # N x J
  P  <- softmax(LP)            # probs
  # dP/dx for each feature k and dest j, averaged over i
  ame <- matrix(NA_real_, nrow = ncol(X_mat), ncol = ncol(B),
                dimnames = list(colnames(X_mat), colnames(B)))
  for (k in seq_len(ncol(X_mat))) {
    # term for each destination j at all i: P_ij * (beta_jk - sum_l P_il * beta_lk)
    beta_k <- B[k, ]                   # length J
    adj    <- P %*% beta_k             # N x 1
    dP     <- P * sweep(matrix(beta_k, nrow(P), ncol(P), byrow=TRUE), 1, adj, FUN = "-")
    ame[k, ] <- colMeans(dP)
  }
  ame
}

# Example: AMEs for one origin
AME <- average_marginal_effects("otherland")
# Rank features by overall AME magnitude across destinations
ame_rank <- data.frame(
  ks = rownames(AME),
  l1_ame = rowSums(abs(AME)),    # or l2_ame = sqrt(rowSums(AME^2))
  stringsAsFactors = FALSE
) |>
  dplyr::arrange(dplyr::desc(l1_ame))
head(ame_rank, 15)

```

### Permutation importance on priors

```{r}
# Build "observed" training matrix for an origin (your Yraw for Ts=2015)
build_Ymat <- function(lu_from) {
  Y <- dplyr::filter(Yraw, lu.from == !!lu_from & Ts == 2015) |>
    tidyr::pivot_wider(names_from = lu.to, values_from = value, values_fill = 0) |>
    tibble::column_to_rownames("ns") |>
    dplyr::select(-lu.from, -Ts)
  # Align rows with X_wide order; fill missing rows with zeros if needed
  Y <- Y[match(rownames(X_mat), rownames(Y)), , drop=FALSE]
  Y[is.na(Y)] <- 0
  as.matrix(Y)
}

logloss_prior <- function(lu_from, X = X_mat) {
  B  <- coef_matrix_for(lu_from)
  P  <- softmax(X %*% B) + 1e-12
  Y  <- build_Ymat(lu_from)
  -sum(Y * log(P))  # cross-entropy
}

perm_importance <- function(lu_from, features = colnames(X_mat), n_perm = 1) {
  base <- logloss_prior(lu_from)
  out <- lapply(features, function(f) {
    ll_inc <- replicate(n_perm, {
      Xp <- X_mat
      Xp[, f] <- sample(Xp[, f])  # permute feature
      logloss_prior(lu_from, X = Xp) - base
    })
    data.frame(ks = f, delta_logloss = mean(ll_inc))
  })
  dplyr::arrange(dplyr::bind_rows(out), dplyr::desc(delta_logloss))
}

# Example: top 15 most important features for origin "otherland"
#perm_importance("otherland") |> head(15)

```

