---
title: "Downscaling"
author: "Clara Douzal"
date: "2023-10-24"
output: pdf_document
params:
  variable1 : country
  variable2 : pathway 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
```

```{r temporary settings}
country = "CHN"
pathway = "CurrentTrends"


```

This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population <- geojson_read(here("Data", country, "Population2020.geojson"),  what = "sp")
population.df <- population@data

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"),  what = "sp")
ProtectedAreas.df <- ProtectedAreas@data

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"),  what = "sp")
Travel.df <- Travel@data

LandCover <- geojson_read(here("Data", country, "LandCoverCopernicus2019.geojson"),  what = "sp")
LandCover.df <- LandCover@data

LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2018_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data

ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")
```

```{r create raster for maps, include = FALSE}
#raster to use for plots
ext <- extent(Travel)
res <- 0.5
raster_layer <- raster(ext, res = res)
rasterized_layer <- rasterize(Travel, raster_layer, field = "fid")
```

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
country_luc <- LandCoverChange.df %>% 
  select(fid) %>% 
  mutate(temp = 1) %>% 
  full_join(map_HILDA%>%  mutate(code = as.character(code)) %>% mutate(temp = 1), relationship = "many-to-many") %>% 
  left_join(LandCoverChange.df %>% 
  pivot_longer(-c(id, fid), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA %>%  mutate(code = as.character(code)), by = c("name" = "code")),
  relationship = "many-to-many") %>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(fid) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(fid, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = fid,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2018) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.to == "forest", "newforest", lu.to)) %>% 
  rbind(data.frame(ns = unique(LandCoverChange.df$fid),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban", "forest"), each = length(unique(LandCoverChange.df$fid))),
                   lu.to = "forest",
                   Ts = 2018,
                   value = 0)) %>% 
  mutate(value = ifelse(lu.from == "forest" & lu.to == "forest", 1, value)) %>% 
  rbind(data.frame(ns = unique(LandCoverChange.df$fid),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), each = length(unique(LandCoverChange.df$fid))),
                   Ts = 2018,
                   value = 0)) 

table(country_luc$lu.from, country_luc$lu.to)
  
```

```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}
#Right now only has land cover data but should contain all the explanatory variables to compute the priors
Xmat <- LandCover.df %>% 
  pivot_longer(-c(id, fid), values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover))  %>% 
  mutate(name = stringr::str_remove(name, "X")) %>% 
  #associate the right land cover names
  left_join(map_Copernicus %>%  mutate(code = as.character(code)), by = c("name" = "code")) %>% 
  group_by(fid, LandCover) %>% 
  #compute the area per cover class in each cell
  summarise(value = sum(AreaPerCover, na.rm = T)) %>% 
  dplyr::select(fid, LandCover, value) %>% 
  #pivot wider to add the intercept value
  pivot_wider(names_from = "LandCover", values_from = "value") %>% 
  mutate(Intercept = 1) %>% 
  #doesn't exist in copernicus and hilda need to add it
  mutate(newforest = 1) %>% 
  #pivot back longer to fit the DownscalR package structure
  pivot_longer(-fid, values_to = "value", names_to = "ks") %>% 
  rename(ns = fid)
```

```{r changing land cover structure to create lu_levels, echo = FALSE, message = FALSE}

#land use level per cell per land cover class

lu_levels <- LandCover.df %>% 
  pivot_longer(-c(id, fid), values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover))  %>% 
  mutate(name = stringr::str_remove(name, "X")) %>% 
  #associate the right land cover names
  left_join(map_Copernicus %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>% 
  filter(!(LandCover %in% c("ocean", "water"))) %>% 
  group_by(fid, LandCover) %>% 
  #Compute the area per cover class per cell (i.e. this is not a share, should it be a share?)
  summarise(value = sum(AreaPerCover, na.rm = T)) %>% 
  rename(ns = fid,
         lu.from = LandCover) %>% 
  dplyr::select(ns, lu.from, value) %>% 
  rbind(data.frame(ns = unique(LandCover.df$fid),
                   lu.from = "newforest",
                   value = 0)) 

```

```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}
country_df <- list(xmat = tibble(Xmat), lu_levels = tibble(lu_levels))
```

# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}
FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"),
                            sheet = pathway) %>% #In calculator, sheet 4_calc_land, 
  #table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  dplyr::select(-YearStart) %>% 
  pivot_longer(-c(YearEnd, LandCoverInit), 
               values_to = "value", 
               names_to = "lu.to") %>% 
  mutate(lu.to = stringr::str_remove(lu.to, "To")) %>% 
  mutate(lu.from = tolower(LandCoverInit),
         lu.to = tolower(lu.to),
         times = YearEnd) %>% 
  dplyr::select(times, lu.from, lu.to, value) %>% 
  #do not keep land cover that remains the same
  filter(!(lu.from == lu.to))
  

table(FABLE$lu.from, FABLE$lu.to)
```

# Estimate MNL model

```{r prep data, echo = FALSE, message=FALSE, include = FALSE}
Yraw <- country_luc 

results_MNL <- data.frame()
pred_coeff_long <- data.frame()
country_start_areas <- data.frame()

for(cur_LU.from in c("urban", "cropland", "pasture", "forest", "otherland", "newforest")){
  
 Y <- dplyr::filter(Yraw,lu.from == cur_LU.from & Ts == 2018) %>% #changes 2000 to 2010
  tidyr::pivot_wider(names_from = lu.to) %>%
  tibble::column_to_rownames(var = "ns") %>%
  dplyr::select(-c(lu.from, Ts))
 
 X <- country_df$xmat %>% 
    tidyr::pivot_wider(names_from = "ks") %>%
    dplyr::arrange(match(ns,Y$ns)) %>%
    tibble::column_to_rownames(var = "ns") 
  
  baseline <- which(colnames(Y) == cur_LU.from)

  # Decreased niter = 100 and nburn = 50 for faster computation 
  temp_results_MNL <- mnlogit(as.matrix(X), as.matrix(Y), baseline, niter = 100, nburn = 50, A0 = 10^4)
  
  pred_coeff <- as.data.frame(apply(temp_results_MNL$postb[,-baseline,],c(1,2),mean))
  
  pred_coeff_long <- pred_coeff_long %>%
    rbind(
    pred_coeff %>% tibble::rownames_to_column(var="ks") %>%
  tidyr::pivot_longer(!ks, names_to = "lu.to", values_to="value") %>%
  dplyr::arrange(lu.to) %>%
  tibble::add_column(lu.from=cur_LU.from, .after="ks")
    )
  
  ## starting area should provide information on area of each (higher resolution) grid cell, i.e ns
country_start_areas <- country_start_areas %>%
  rbind(
    dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>% mutate(times = 2020)
  )

}
```

```{r, echo = FALSE, message=FALSE}
X_long <-  X %>% tibble::rownames_to_column(var="ns") %>%
  tidyr::pivot_longer(!ns, names_to="ks", values_to = "value")

## Downscaling computation and results

results_DS <- downscale(targets = FABLE, 
                        start.areas =  country_start_areas, 
                        xmat = X_long, 
                        betas = pred_coeff_long)

downscaled_LUC <- results_DS$out.res

```

# Land-use map

```{r, echo = FALSE, message=FALSE}
LUC_plot <- LUC_plot(results_DS, rasterized_layer)
LUC_plot$LUC.plot 


```

# Land-use change map

```{r}

# Need to tweak it a bit to make sure that we have loss and gains, this is not the case right now (27/10/2023)


results_DS_LUC <- results_DS
results_DS_LUC$out.res <- results_DS_LUC$out.res %>% 
  #remove the stable transition
  filter(!(lu.from == lu.to)) %>% 
  #compute the loss and gains within stable land cover
  rbind(
    results_DS_LUC$out.res %>% 
    filter(lu.from == lu.to) %>% 
    arrange(lu.to, ns, times) %>% 
    group_by(lu.to, ns) %>% 
    mutate(value = value - lag(value))
  ) %>% 
  #remove first time periode for stable transition
  filter(!is.na(value))

color = "PRGn"
label = "Area gains and losses in ha per pixel"
Transition_plot <- LUC_plot(results_DS_LUC, rasterized_layer, 
                            color = "PRGn",
                            label = "Area gains and losses in ha per pixel")

limit <- max(abs(Transition_plot[["plot.df"]]$value)) * c(-1, 1)
Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2()
#palette = color, 
                                #name=label)#, 
                                #direction = 1)#, 
                                #limit = limit)


plot_obj <- ggplot2::ggplot() +
    ggplot2::geom_tile(data=Transition_plot[["plot.df"]], 
                       aes(x=x, y=y, 
                           # fill=factor(cut(value, breaks =  unique(quantile(value, probs = seq(0.,1,0.10))))),
                           fill = factor(cut(value, 
                                             breaks = fullseq(range(value),(range(value)[2]-range(value)[1])/8))),
                           group=lu.to), 
                       alpha=0.8) +
    ggplot2::scale_fill_manual(values = RColorBrewer::brewer.pal(9, "PRGn"),#palette = "PRGn",
                                  #direction = 1,
                                  #limit = limit#,
                                  #name=label
                                  name = NULL)+
    ggplot2::coord_equal() +
    ggthemes::theme_map() +
    ggplot2::theme(legend.position="bottom") +
    ggplot2::theme(legend.key.width=unit(2, "cm"))+
    ggplot2::facet_grid(times~lu.to)
  
plot_obj

```
