---
title: "Downscaling"
author: "Clara Douzal"
date: "`r format(Sys.time(), '%y%m%d %H:%M')`"
output: pdf_document
header-includes:
  - \DeclareUnicodeCharacter{2061}{}
  - \DeclareUnicodeCharacter{2212}{}
params:
  country : "BRA"
  pathway : "CurrentTrends"
  start_map_source : "HILDA"
  mnl_niter : 100
  mnl_nburn : 50
  seed : 1212
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 150)

```



```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
library(stringr)
#library(kableExtra)
library(knitr)
library(readr)
library(forcats)
library(purrr)
library(writexl)
library(ggplot2)

if (requireNamespace("conflicted", quietly = TRUE)) {
  conflicted::conflict_prefer("filter",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("select",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("lag",       "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("rename",    "dplyr", quiet = TRUE)
  conflicted::conflict_prefer("left_join", "dplyr", quiet = TRUE)
}
```

```{r setting paths, include = FALSE}
set.seed(params$seed)
country <- params$country
pathway <- params$pathway
niter   <- params$mnl_niter
nburn   <- params$mnl_nburn
start_map_source <- params$start_map_source

dir.create(here("Output", country), showWarnings = FALSE, recursive = TRUE)
stamp <- format(Sys.Date(), "%y%m%d")
tag   <- paste(stamp, country, pathway, start_map_source, sep = "_")

```

```{r loading special functions, include = FALSE}
source(here("Program", "LUC_plot_cum_allLUC.R"))
source(here("Program", "LUC_plot_compare_pathways.R"))
source(here("Program", "wrangle_plot.R"))
```
 

This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE, cache = FALSE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population    <- geojson_read(here("Data", country, "Population2020.geojson"), what = "sp")
population.df <- population@data %>%
  select(-id) #%>%
#mutate(TotalPop = c(scale(TotalPop, scale = T)))

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"), what = "sp")
ProtectedAreas.df <- ProtectedAreas@data %>%
  select(-id)

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"), what = "sp")
Travel.df <- Travel@data %>%
  select(-id)


if(start_map_source == "HILDA"){
#this is in km2 
LandCover <- geojson_read(here("Data", country, "LandCoverHILDA2015.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)
} else {
  LandCover <- geojson_read(here("Data", country, "LandCoverCopernicus2019.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)
} 

#this is in km2
LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2015_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data %>% 
   select(-id)


ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data %>% 
  select(-id)

altitude <- geojson_read(here("Data", country, "Altitude.geojson"),  what = "sp")
altitude.df <- altitude@data %>% 
  select(-id)

slope <- geojson_read(here("Data", country, "Slope.geojson"),  what = "sp")
slope.df <- slope@data %>% 
  select(-id)

gaez <- geojson_read(here("Data", country, "GAEZCropDistribution2015.geojson"),  what = "sp")
gaez.df <- gaez@data  %>% 
  select(-id) %>% 
  select(-total)

livestock <- geojson_read(here("Data", country, "GLW4WorldGriddedLivestock2020.geojson"),  what = "sp")
livestock.df <- livestock@data  %>% 
  select(-id) %>% 
  select(-alllivestock)

#Loading the 50km grid with
grid <- read_csv(here("Data", "global", "grid50_equal_area.csv")) %>% 
  filter(iso3 == country)

zones <- grid %>% 
  select(id_c, thermal, zone, BIOME_NAME)

# ab_be_biomass <- geojson_read(here("Data", country, "AbovegroundBelowgroundBiomassCarbonDensity.geojson"),  what = "sp")
# ab_be_biomass.df <- ab_be_biomass@data %>% 
#   select(-id)

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_HILDA_LUC <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA_change")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")

map_ForestMgmt <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ForestManagement")

map_protectedareas <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ProtectedAreas")
```


```{r restoration maps, include = FALSE}

# #mapping
# map_chaturvedi <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
#                                 sheet = "Chaturvedi")
# 
# #Global tree cover (existing and potential) (%) (remove existing? which map is used for existing?)
# rest_bastin <- geojson_read(here("Data", country, "GlobalTreeRestorationPotential_Bastin.geojson"),  what = "sp")
# rest_bastin.df <- rest_bastin@data  %>% 
#   filter(!is.na(mean)) %>% 
#   left_join(grid %>% select(id_c, area)) %>% 
#    mutate(rest_area = pmin(area, (mean/100)*area)) %>% 
#   select(-id) 
# 
# #kilograms of CO2 equivalent per year (kgCO2e yr-1)
# #transformed into t CO2/ 1000 ha
# rest_griscom <- geojson_read(here("Data", country, "ReforestationPotential_Griscom.geojson"),  what = "sp")
# rest_griscom.df <- rest_griscom@data  %>% 
#   select(-id) %>% 
#   left_join(grid %>% select(id_c, area)) %>% 
#   mutate(p = (sum / 1000)/(1000 *area)) #p in t CO2/ 1000 ha
# 
# #absolute hectare per cells
# #transformed into percentage of restoration potential type per cell
# if(country == "IND"){
#   rest_chaturvedi <- geojson_read(here("Data", country, "LandscapeRestorationOpportunities.geojson"),  
#                                   what = "sp")
#  rest_chaturvedi.df <- rest_chaturvedi@data  %>% 
#   select(-id) %>% 
#   # %>% 
#   pivot_longer(-c(id_c)) %>% 
#    mutate(code = stringr::str_remove(name, "X")) %>%
#   left_join(map_chaturvedi %>% mutate(code = as.character(code))) %>% 
#    left_join(grid %>% select(id_c, area)) %>% 
#    mutate(p = pmin(100, 100 * value/area))
# }
# 
# #Area with potential for forest regeneration per pixel
# #Values are given in hectares
# rest_williams <- geojson_read(here("Data", country, "PotentialForestRegenerationTropicalForest_Williams.geojson"),  
#                                   what = "sp")
# rest_williams.df <- rest_williams@data  %>% 
#   select(-id) 
# 
# #Constrained unrealized potential aboveground biomass, belowground biomass, and soil organic carbon combined density under baseline climate (Mg C ha-1
# #Values are given in Mg C
# rest_walker <- geojson_read(here("Data", country, "UnrealizedPotentialIncreaseStorageCarbonLand_Walker.geojson"),  
#                                   what = "sp")
# rest_walker.df <- rest_walker@data  %>% 
#   select(-id) 
# 
# #Net Climate Impact of tree cover restoration (carbon storage and albedo change) (Mg CO2e ha-1)
# #Values are given in Mg CO2 equivalent
# 
# #Fesenmyer - scenario based restoration potential area in hectares
# rest_fesenmyer <- geojson_read(here("Data", country, "RestorationFesenmyer.geojson"),  
#                                   what = "sp")
# rest_fesenmyer.df <- rest_fesenmyer@data %>% 
#   select(id_c, d0)



```


Below is the table of input data used in the prior module of the downscaling process.

```{r echo=FALSE, message=FALSE}
library(knitr)

# Create the data frame for the table
table_data <- data.frame(
  `Variable in econometric model` = c("Dependent", "Land use change", "", 
                                      "Explanatory variables", "Land use", 
                                      "Forest Management", "Mean Travel time to major cities", "Population", "Crop yield (27 indicidual crops)"),
  Year = c("", "2015-2019", "", "", "2015", "2015", "2000", "2020", "2015"),
  Unit = c("", "% of lu.from in pixel being converted", "", "", 
           "hectares", "hectares", "minutes", "Count", "kg/ha"),
  Source = c("", "HILDA", "", "", "COPERNICUS", "Lesiv et al.", 
             "Nelson, A (2008)", "WorldPop", "GAEZ+")
)

# Use kable without booktabs
kable(
  table_data, 
  format = "latex", 
  col.names = c("Variable in econometric model", "Year", "Unit", "Source")
)
```


## Checking the land cover change input data

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix. With "missing_from" you can see the number of land transitions thatare missing because of a missing initial land cover. With "missing_to" you can see the number of land transitions that are missing because of a missing final land cover.


```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
#value is the percentage of the cell that is going from land cover to another
country_luc <- LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA_LUC %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water", "not relevant"))) %>% 
  filter(!(to %in% c("ocean", "water", "not relevant"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2015) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2015,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) 

# table(country_luc$lu.from, country_luc$lu.to)
  
expected <- c("cropland","forest","newforest","otherland","pasture","urban")
present_from <- sort(unique(country_luc$lu.from))
present_to   <- sort(unique(country_luc$lu.to))
data.frame(kind = c("missing_from","missing_to"),
           value = I(list(setdiff(expected, present_from),
                         setdiff(expected, present_to)))) %>% knitr::kable()
```

If any land transitions are missing you can check the table below.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r , echo = FALSE, message = FALSE}
table(country_luc$lu.from, country_luc$lu.to) %>% knitr::kable()
```

# Harmonizing start map with FABLE Calculator baseline map

We need to harmonize the high-resolution baseline map used in the downscalR function with the aggregated land-use data from the FABLE Calculator. This ensures that the spatially explicit baseline map aligns with the aggregated land-use data in FABLE. Without this alignment, there could be discrepancies in the starting levels of each land-use class, making the future land-use transitions projected by the FABLE Calculator unrealistic when applied to the unadjusted high-resolution map.

```{r changing land cover structure to create start_map harmonized with FABLE aggregated land cover, echo = FALSE, message = FALSE}

#land use level per cell per land cover class in baseline year
start_map <- LandCover.df %>%
  # Pivot to long format and convert AreaPerCover to numeric
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_Copernicus and filter out "ocean" and "water"
  left_join(map_HILDA %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  filter(!LandCover %in% c("ocean", "water")) %>%
  # Group by id_c and LandCover, summarizing the total area per cover class
  group_by(id_c, LandCover) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha - 1 km² = 0.1 thousand ha
  summarise(value = sum(AreaPerCover, na.rm = TRUE)/10, .groups = 'drop') %>%
  # Rename columns for clarity
  rename(ns = id_c, lu.from = LandCover) %>%
  # Add rows for "newforest" with value = 0 for each unique id_c
  bind_rows(data.frame(ns = unique(LandCover.df$id_c), lu.from = "newforest", value = 0)) %>% 
  mutate(lu.from = ifelse(is.na(lu.from), "not relevant", lu.from))
  
start_map_harm <- start_map %>% 
  group_by(lu.from) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(lu = lu.from) %>% 
  mutate(lu = ifelse(is.na(lu), "not relevant", lu))

#FABLE Calculator aggregated land use level in baseline year
LC_targets <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = "Baseline") %>% 
  rename(lu = LandCover)

keep <- "from" # Specify which areas to keep during transition

# Calculate land-use change targets based on the start and target data
LUC_targets <- LU_to_LUC(start_map_harm, 
                              LC_targets, 
                              keep_areas = keep)

# Create a check data frame to validate the results
check_df <- rbind(NULL,
                  left_join(
                    start_map_harm %>% group_by(lu) %>% 
                      rename(lu.from = "lu") %>% 
                      reframe(value_start = sum(value)), 
                    LC_targets %>% group_by(lu) %>% 
                      rename(lu.to = "lu") %>% 
                      reframe(value_target = sum(value)), 
                    by = c("lu.from" = "lu.to")) %>%
                    left_join(LUC_targets %>% 
                                group_by(lu.to) %>% 
                                rename(value_new = "value") %>% 
                                reframe(value_new = sum(value_new)), 
                              by = c("lu.from" = "lu.to")) %>% 
                    select(lu.from, value_start, value_target, value_new) %>%
                    mutate(new_target_diff = value_new - value_target, 
                           value_rebalance = sum(value_start) * (value_target / sum(value_target)), 
                           rebalance_new_diff = value_rebalance - value_new))

### Calculate implied transition shares from LUC targets
CLC_FABLE_shares <- LUC_targets %>% 
  group_by(lu.from) %>% 
  mutate(value = value / sum(value)) %>% # Normalize values to get shares
  pivot_wider(id_cols = c(lu.from), names_from = "lu.to", values_from = "value") %>% 
  replace(is.na(.), 0) %>% # Replace NA with 0 for clarity
  pivot_longer(cols = !c(lu.from), names_to = "lu.to", values_to = "proj_share")

# Apply the transition shares to the high-resolution start map for downscaling
start_map_reproj <- left_join(start_map, CLC_FABLE_shares, relationship = "many-to-many") %>% 
  mutate(value = ifelse(!is.na(lu.to), value * proj_share, value), # Adjust values based on shares
         lu.to = ifelse(is.na(lu.to), lu.from, lu.to)) %>% # Update land use type
  group_by(ns, lu.to) %>% 
  summarize(value = sum(value)) %>% # Sum values for each unique combination
  ungroup() %>% 
  drop_na() %>% # Remove rows with NA values
  rename(lu.from = "lu.to") # Rename for consistency

saveRDS(start_map_reproj, here("Output", country, paste0(tag, "_start_map_reproj.rds")))
saveRDS(start_map, here("Output", country, paste0(tag, "_start_map.rds")))

```

```{r create a stable ID map, include = FALSE}

# === build a master ID map (character id -> integer code) ===
# start_map already has ns (from id_c)
ns_map <- start_map %>%
  dplyr::distinct(ns) %>%
  dplyr::mutate(ns = as.character(ns)) %>%
  dplyr::arrange(ns) %>%
  dplyr::mutate(ns_int = dplyr::row_number(),
                id_c   = ns)  # convenience column for joining to sp@data$id_c

saveRDS(ns_map, here("Output", country, paste0(tag, "_ns_map.rds")))
# 
# 
# # sanity: no duplicates
# stopifnot(!anyDuplicated(ns_map$ns))
# 
# # === rasterize *now* (doesn't need results yet) ===
# # make sure Travel has id_c and same CRS
# # Travel@data <- Travel@data %>%
# #   dplyr::mutate(id_c = as.character(id_c)) %>%
# #   dplyr::left_join(ns_map, by = "id_c")
# 
# # fail fast if some polygons aren't in the map
# if (anyNA(Travel@data$ns_int)) {
#   missing <- unique(Travel@data$id_c[is.na(Travel@data$ns_int)])
#   stop(sprintf("Travel polygons not found in ns_map (n=%d). Example: %s",
#                length(missing), paste(head(missing, 5), collapse=", ")))
# }
# 
# raster_layer <- raster::raster(raster::extent(Travel), res = 0.5, crs = raster::crs(Travel))
# rasterized_layer <- raster::rasterize(Travel, raster_layer, field = "ns_int")
# 
# saveRDS(rasterized_layer, here("Output", country, paste0(tag, "_rasterized_layer.rds")))


```

```{r}
make_id_raster <- function(Travel_sp, ns_map,
                           keep_ns_int,
                           res_km = 5) {

  # --- clean ns_map ---
  ns_map <- ns_map %>%
    dplyr::mutate(
      id_c   = as.character(id_c),
      ns_int = as.integer(ns_int)
    ) %>%
    dplyr::distinct(id_c, ns_int)

  # --- attach ns_int to Travel ---
  Travel_sp@data <- as.data.frame(Travel_sp@data)
  Travel_sp@data$id_c <- as.character(Travel_sp@data$id_c)

  Travel_sp@data <- Travel_sp@data %>%
    dplyr::select(-dplyr::any_of(c("ns_int","ns_int.x","ns_int.y"))) %>%
    dplyr::left_join(ns_map, by = "id_c") %>%
    as.data.frame()

  # --- keep only valid cells ---
  Travel_sp <- Travel_sp[Travel_sp@data$ns_int %in% keep_ns_int, ]
  Travel_sp@data <- as.data.frame(Travel_sp@data)

  # --- rasterize with terra ---
  v <- terra::vect(Travel_sp)
  v <- terra::project(v, "EPSG:6933")   # equal-area

  r <- terra::rast(
    terra::ext(v),
    resolution = res_km * 1000,
    crs = terra::crs(v)
  )

  terra::rasterize(
    v, r,
    field   = "ns_int",
    touches = TRUE
  )
}

# keep only cells with valid travel time
keep_id_c <- Travel@data %>%
  dplyr::mutate(id_c = as.character(id_c)) %>%
  dplyr::filter(!is.na(MeanTravelTime)) %>%
  dplyr::pull(id_c) %>%
  unique()

keep_ns_int <- ns_map %>%
  dplyr::mutate(id_c = as.character(id_c)) %>%
  dplyr::filter(id_c %in% keep_id_c) %>%
  dplyr::pull(ns_int) %>%
  unique()

rasterized_layer <- make_id_raster(
  Travel,
  ns_map,
  keep_ns_int = keep_ns_int,
  res_km = 5
)

# df_map <- terra::as.data.frame(rasterized_layer, xy = TRUE, na.rm = FALSE)
# names(df_map)[3] <- "ns_int"

terra::writeRaster(
  rasterized_layer,
  here("Output", country, paste0(tag, "_rasterized_layer.tif")),
  overwrite = TRUE
)


grid_sf <- sf::st_as_sf(Travel) %>%
  dplyr::select(id_c) %>%        # keep only ID
  sf::st_make_valid()

saveRDS(grid_sf, here("Output", country, paste0(tag, "_grid_sf.rds")))
```


# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}

FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = pathway) %>%
  # In calculator, sheet 4_calc_land, 
  # table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  
  # Remove YearStart and reshape to long format
  select(-YearStart) %>%
  pivot_longer(-c(YearEnd, LandCoverInit), values_to = "value", names_to = "lu.to") %>%
  # Clean and adjust land cover names
  mutate(
    lu.from = tolower(LandCoverInit),
    lu.to = tolower(str_remove(lu.to, "To")),
    times = YearEnd
  ) %>%
  # Select necessary columns and filter out unchanged land covers
  select(times, lu.from, lu.to, value) #%>%
  #filter(lu.from != lu.to)
  
FABLE <- FABLE %>%
  # Remap non-forest -> forest to newforest (afforestation)
  mutate(lu.to = if_else(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>%
  # Drop impossible pairs:
  #  - forest -> newforest (not modeled)
  #  - newforest -> anything (no outflows from newly afforested land)
  filter(!(lu.from == "forest"   & lu.to == "newforest")) %>%
  filter(!(lu.from == "newforest")) %>%
  # Collapse duplicates created by the remap; keep off-diagonal only
  group_by(times, lu.from, lu.to) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") #%>%
  #filter(lu.from != lu.to)

table(FABLE$lu.from, FABLE$lu.to) %>% knitr::kable()

saveRDS(FABLE %>% dplyr::filter(times >= 2020), here("Output", country, paste0(tag, "_FABLE.rds")))



```


```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}

Xmat_LU <- start_map_reproj %>% 
  pivot_wider(names_from = "lu.from", values_from = "value", values_fill = 0) %>% 
  rename(id_c = ns)

#ForestManagement
Xmat_ForestManagement <- ForestManagement.df %>%
  # Pivot longer to convert cover columns to rows
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_ForestMgmt to get the corresponding forest management types
  left_join(map_ForestMgmt %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  # Group by id_c and type to summarize total area per cover class per pixel
  group_by(id_c, type) %>%
  summarise(value = sum(AreaPerCover, na.rm = TRUE), .groups = 'drop') %>%
  # Pivot wider to create columns for each forest management type
  pivot_wider(names_from = type, values_from = value, values_fill = 0) 


#Mean Travel time
Xmat_Travel <- Travel.df %>% select(id_c, MeanTravelTime) %>% 
  transmute(id_c, log_MeanTravelTime = log1p(MeanTravelTime)) 


#Population  
Xmat_pop <- population.df %>% select(id_c, TotalPop) %>% 
  transmute(id_c, log_pop = log1p(TotalPop)) 
 
  
#Altitude  
Xmat_alt <- altitude.df %>% select(id_c, MeanAltitude) %>% 
  transmute(id_c, log_altitude = log1p(MeanAltitude))

Xmat_slope <- slope.df %>%
  dplyr::transmute(
    id_c     = as.character(id_c),
    log_slope = log1p(MeanSlope))

Xmat_zones <- zones %>% 
  dplyr::transmute(
    id_c     = as.character(id_c),
    thermal  = forcats::fct_explicit_na(as.factor(thermal), na_level = "UNK"),
    zone     = forcats::fct_explicit_na(as.factor(zone),    na_level = "UNK")
  ) %>%
  { 
    MM <- model.matrix(~ thermal + zone, data = .)   # has (Intercept) + k-1 dummies
    MM <- MM[, setdiff(colnames(MM), "(Intercept)"), drop = FALSE]  # drop intercept
    cbind(id_c = .$id_c, as.data.frame(MM))
  } %>%
  tibble::as_tibble() %>%
  # full coverage of pixels
  #dplyr::right_join(tibble::tibble(id_c = all_ns), by = "id_c") %>%
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("thermal"), ~ tidyr::replace_na(.x, 0)),
    dplyr::across(dplyr::starts_with("zone"),    ~ tidyr::replace_na(.x, 0))
  )

#### Crop yield data #####

# gaez_yield.df: id_c + 25 crop yields (kg/ha)
yield_cols <- setdiff(names(gaez.df), "id_c")

# 1) log-transform (handles skew and zeros)
Ylog <- gaez.df  %>% 
  dplyr::mutate(dplyr::across(all_of(yield_cols), ~ log1p(pmax(.x, 0))))

# 2) overall level per cell (mean log-yield across crops), then z-score
Xmat_yield_level <- Ylog  %>% 
  dplyr::rowwise()  %>% 
  dplyr::mutate(level_mean_log = mean(c_across(all_of(yield_cols)), na.rm = TRUE))  %>% 
  dplyr::ungroup()  %>% 
  dplyr::transmute(
    id_c,
    m = m <- mean(level_mean_log, na.rm=TRUE),
    s = sd(level_mean_log, na.rm=TRUE),
    level = level_mean_log,
    yield_level_z = ifelse(s>0, (level_mean_log - m)/s, 0) 
  ) %>% 
  select(id_c, yield_level_z)

# Compute per-cell relative advantage from crop yields
# - Row-center (relative to the cell's mean across crops)
# - Optional log1p transform (good for skewed yields & zeros)
# - Column z-score (so coefficients are comparable)
# - Drops near-constant crops automatically
reladv_yields <- function(df, id_col = "id_c", crop_cols = NULL,
                          transform = c("log1p", "none"),
                          winsor_p = NULL,          # e.g., 0.01 to tame outliers
                          nzv_thresh = 1e-6) {
  
  stopifnot(id_col %in% names(df))
  transform <- match.arg(transform)
  
  # pick crop columns
  if (is.null(crop_cols)) {
    crop_cols <- setdiff(names(df), id_col)
  }
  X <- as.matrix(df[, crop_cols, drop = FALSE])
  
  # NAs -> 0 (no production) before transforms
  X[is.na(X)] <- 0
  
  # optional winsorization per column - cap extreme values at chosen quantiles
  # Reduces the influence of extreme points on models/means/SDs while keeping sample size.
  if (!is.null(winsor_p) && winsor_p > 0 && winsor_p < 0.5) {
    qlo <- apply(X, 2, quantile, probs = winsor_p,      na.rm = TRUE)
    qhi <- apply(X, 2, quantile, probs = 1 - winsor_p,  na.rm = TRUE)
    X   <- pmin(pmax(X, rep(qlo, each = nrow(X))), rep(qhi, each = nrow(X)))
  }
  
  # variance-stabilize (good for yields with zeros)
  if (transform == "log1p") {
    X <- log1p(pmax(X, 0))
  }
  
  # --- Relative advantage: subtract per-row mean across crops ---
  row_mu <- rowMeans(X)                   # N x 1
  Xc     <- sweep(X, 1, row_mu, "-")      # N x C (row-centered)
  
  # --- Column wise standardization & NZV drop ---
  col_sd <- apply(Xc, 2, sd)
  keep   <- which(is.finite(col_sd) & col_sd >= nzv_thresh)
  if (length(keep) == 0L) {
    stop("All relative-advantage crop columns are (near) constant; check inputs.")
  }
  
  # center & scale by column (guard tiny sds)
  col_mu_keep <- colMeans(Xc[, keep, drop = FALSE])
  col_sd_keep <- pmax(apply(Xc[, keep, drop = FALSE], 2, sd), 1e-12)
  Xcz <- sweep(sweep(Xc[, keep, drop = FALSE], 2, col_mu_keep, "-"),
               2, col_sd_keep, "/")
  
  # back to tibble: id + kept crops
  out <- tibble::tibble(!!id_col := df[[id_col]]) %>%
    dplyr::bind_cols(as.data.frame(Xcz))
  names(out)[-1] <- colnames(Xc)[keep]
  out
}

# Your yield table: id_c + 25 crop yield columns (kg/ha)
yield_cols <- setdiff(names(gaez.df), "id_c")

Xmat_yield_reladv <- reladv_yields(
  df         = gaez.df,
  id_col     = "id_c",
  crop_cols  = yield_cols,
  transform  = "log1p",   # good default
  winsor_p   = NULL       # or 0.01 if you have wild outliers
)



Xmat_livestock <- livestock.df %>% 
  mutate(id_c     = as.character(id_c)) %>% 
  #drop columns that are all zeros
  select(
    any_of("id_c"),                               # keep id_c if present
    where(~ !is.numeric(.x) || any(abs(.x) > 1e-9, na.rm = TRUE))  # keep non-numeric OR has any non-zero
  )
# see which columns were dropped
# dropped_cols <- setdiff(names(livestock.df), names(Xmat_livestock))
# print(dropped_cols)

Xmat <- Xmat_LU %>%
  mutate(Intercept = 1) %>%
  left_join(Xmat_Travel,           by = "id_c") %>%
  left_join(Xmat_pop,              by = "id_c") %>%
  left_join(Xmat_ForestManagement, by = "id_c") %>%
  left_join(Xmat_alt,              by = "id_c") %>%
  left_join(Xmat_slope,            by = "id_c") %>%
  left_join(Xmat_zones,            by = "id_c") %>% 
  left_join(Xmat_yield_reladv,     by = "id_c") %>% 
  left_join(Xmat_yield_level,      by = "id_c") %>% 
  left_join(Xmat_livestock,        by = "id_c") %>% 
  pivot_longer(-id_c, names_to = "ks", values_to = "value") %>%
  rename(ns = id_c)


```

We cannot keep the cells that have NAs for some variables that will be used as priors.

```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}

# After you build Xmat (long: ns, ks, value)
na_by_var <- Xmat %>%
  group_by(ks) %>%
  summarise(n_na = sum(is.na(value)), .groups="drop") %>%
  arrange(desc(n_na))

na_by_ns <- Xmat %>%
  group_by(ns) %>%
  summarise(n_na = sum(is.na(value)), .groups="drop") %>%
  arrange(desc(n_na))

head(na_by_var, 40) %>% knitr::kable()# which variables have NAs?
head(na_by_ns, 20) %>% knitr::kable()   # which cells have most NAs?


good_ns <- Xmat %>% 
  group_by(ns) %>% 
  summarise(ok = all(!is.na(value))) %>% 
  filter(ok) %>%
  pull(ns)

# Which cells were removed?
all_ns <- sort(unique(Xmat$ns))
removed_ns <- setdiff(all_ns, good_ns)

removal_msg <- if (length(removed_ns) == 0) {
  sprintf("**No grid cells were removed**: all %d cells have complete covariates.", length(all_ns))
} else {
  ids <- paste(head(removed_ns, 20), collapse = ", ")
  extra <- if (length(removed_ns) > 20) " ..." else ""
  sprintf("**%d grid cell(s) were removed** due to missing covariates. Removed cell IDs (showing up to 20): %s%s",
          length(removed_ns), ids, extra)
}

country_df <- list(
  xmat = Xmat %>% filter(ns %in% good_ns),
  lu_levels = start_map_reproj %>% filter(ns %in% good_ns)
)


saveRDS(country_df, here("Output", country, paste0(tag, "_countrydf.rds")))
```
`r removal_msg`

# Estimate MNL model


```{r prep data, echo = FALSE, message=FALSE, include = FALSE, cache = FALSE}
Yraw <- country_luc %>% 
  filter(ns %in% good_ns) 

# 1) ONE global wide X for modeling (all cells, all ks)
X_wide <- country_df$xmat %>%
  tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
  dplyr::arrange(ns) %>%
  tibble::column_to_rownames(var = "ns")

# Optional safety check: no NAs
stopifnot(!any(is.na(X_wide)))

results_MNL         <- data.frame()
pred_coeff_long     <- data.frame()
country_start_areas <- data.frame()


# --- build a cleaned, wide design once (drop constants; optional scaling) ---
X_wide0 <- country_df$xmat %>%
  tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
  dplyr::arrange(ns)

# separate ns and features
feat_names <- setdiff(colnames(X_wide0), "ns")
X0         <- as.matrix(X_wide0[, feat_names, drop = FALSE])

# drop constant columns globally
sd_all     <- apply(X0, 2, sd)
keep_feat  <- names(sd_all)[sd_all > 0]
X0         <- X0[, keep_feat, drop = FALSE]
feat_names <- colnames(X0)

# (optional) mild standardization for conditioning (skips columns with tiny sd)
mu         <- colMeans(X0)
sdv        <- apply(X0, 2, sd)
sdv[sdv < 1e-12] <- 1  # guard
X0         <- sweep(sweep(X0, 2, mu, "-"), 2, sdv, "/")

rownames(X0) <- X_wide0$ns  # keep row keys

# containers
pred_coeff_long <- dplyr::tibble()
country_start_areas <- dplyr::tibble()

# helper: keep independent columns via QR (removes linear combos per-origin)
keep_independent_cols <- function(M, tol = 1e-10) {
  q <- qr(M, tol = tol, LAPACK = TRUE)
  M[, q$pivot[seq_len(q$rank)], drop = FALSE]
}

# helper: choose a valid baseline (prefer self; else the destination with largest total)
choose_baseline <- function(Ymat, origin) {
  if (origin %in% colnames(Ymat)) {
    which(colnames(Ymat) == origin)
  } else {
    which.max(colSums(Ymat))
  }
}

# loop over origins with guards + ridge fallback grid
origins <- c("urban","cropland","pasture","forest","otherland","newforest")
ridge_grid <- c(10, 50, 100, 500)  # stronger shrinkage = more stable

for (cur_LU.from in origins) {

  # build Y for this origin and align to X rows
  Yfull <- dplyr::filter(Yraw, lu.from == cur_LU.from & Ts == 2015) %>%
    tidyr::pivot_wider(names_from = lu.to, values_from = value, values_fill = 0) %>%
    tibble::column_to_rownames("ns") %>%
    dplyr::select(-lu.from, -Ts)

  # align order to X0
  Yfull <- Yfull[match(rownames(X0), rownames(Yfull)), , drop = FALSE]
  Yfull[is.na(Yfull)] <- 0

  # drop destination columns that are all zero
  # keep_dest <- which(colSums(Yfull) > 0)
  # if (length(keep_dest) < 2L) {
  #   message("Skipping origin '", cur_LU.from, "': fewer than 2 non-zero destinations.")
  #   next
  # }
  # Y <- as.matrix(Yfull[, keep_dest, drop = FALSE])

  # subset X rows and drop collinear columns for this subset
  X <- X0[rownames(Y), , drop = FALSE]
  # drop zero-variance on this subset (can re-appear after subsetting)
  sd_sub <- apply(X, 2, sd)
  X <- X[, sd_sub > 0, drop = FALSE]
  # drop linear combinations (QR)
  X <- keep_independent_cols(X)

  # choose a robust baseline
  baseline <- choose_baseline(Y, cur_LU.from)

  # try fitting with increasing ridge until it works
  fit_ok <- FALSE
  for (A0 in ridge_grid) {
    tmp <- try(
      mnlogit(X, Y, baseline, niter = niter, nburn = nburn, A0 = A0),
      silent = TRUE
    )
    if (!inherits(tmp, "try-error")) { fit_ok <- TRUE; break }
  }
  if (!fit_ok) {
    message("mnlogit failed for origin '", cur_LU.from, "' even with ridge; continuing.")
    next
  }

  # posterior means -> long coef table
  post <- tmp$postb
  # drop baseline column in postb
  Bmean <- as.data.frame(apply(post[, ,#-baseline, 
                                    , drop = FALSE], c(1,2), mean))
  coef_long <- Bmean %>%
    tibble::rownames_to_column("ks") %>%
    tidyr::pivot_longer(-ks, names_to = "lu.to", values_to = "value") %>%
    dplyr::mutate(lu.from = cur_LU.from, .before = 1)

  pred_coeff_long <- dplyr::bind_rows(pred_coeff_long, coef_long)

  # accumulate start areas for this origin
  country_start_areas <- dplyr::bind_rows(
    country_start_areas,
    dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>%
      dplyr::mutate(times = 2020)
  )
}

# Which (lu.from, lu.to) target pairs need priors?
target_pairs <- FABLE %>%
  dplyr::filter(times > 2020, value > 0) %>%
  dplyr::distinct(lu.from, lu.to)

# Which pairs do we have betas for?
beta_pairs <- pred_coeff_long %>%
  dplyr::distinct(lu.from, lu.to)

# Pairs missing in betas:
missing_pairs <- dplyr::anti_join(target_pairs, beta_pairs, by = c("lu.from","lu.to"))

if (nrow(missing_pairs) > 0) {
  message("Completing betas with flat priors for missing pairs: ",
          paste(paste0(missing_pairs$lu.from, "->", missing_pairs$lu.to), collapse = ", "))

  ks_kept <- sort(unique(pred_coeff_long$ks))           # features used in betas
  filler <- tidyr::crossing(missing_pairs, ks = ks_kept) %>%
    dplyr::mutate(value = 0)

  pred_coeff_long <- dplyr::bind_rows(pred_coeff_long, filler)
}

# the features that appear in your estimated betas (union across origins)
keep_ks <- sort(unique(pred_coeff_long$ks))
stopifnot(length(keep_ks) > 0)

# build the same wide matrix you used pre-loop
X_wide0 <- country_df$xmat %>%
  tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
  dplyr::arrange(ns)

# means/sds only for the kept features (guard tiny sds)
mu  <- vapply(keep_ks, function(k) mean(X_wide0[[k]], na.rm = TRUE), numeric(1))
sdv <- vapply(keep_ks, function(k)  sd (X_wide0[[k]], na.rm = TRUE), numeric(1))
sdv[is.na(sdv) | sdv < 1e-12] <- 1  # protect against constants

X_long <- country_df$xmat %>%
  dplyr::filter(ks %in% keep_ks) %>%
  dplyr::mutate(
    # apply the same centering/scaling used in the MNL fit
    value = (value - mu[ks]) / sdv[ks]
  ) %>%
  dplyr::arrange(ns, ks)

ks_pred <- sort(unique(pred_coeff_long$ks))
ks_x    <- sort(unique(X_long$ks))
if (!setequal(ks_pred, ks_x)) {
  message("In betas not in X_long: ", paste(setdiff(ks_pred, ks_x), collapse=", "))
  message("In X_long not in betas: ", paste(setdiff(ks_x, ks_pred), collapse=", "))
}
stopifnot(setequal(ks_pred, ks_x))

# Safety: dimensions align
stopifnot(setequal(unique(pred_coeff_long$ks), unique(X_long$ks)))

results_DS <- downscale(
  targets     = FABLE %>% dplyr::filter(times >= 2020),
  start.areas = country_start_areas,
  xmat        = X_long,
  betas       = pred_coeff_long
)

saveRDS(country_start_areas, here("Output", country, paste0(stamp, "_country_start_areas.rds")))
saveRDS(X_long, here("Output", country, paste0(stamp, "_X_long.rds")))
saveRDS(pred_coeff_long, here("Output", country, paste0(stamp, "_pred_coeff_long.rds")))
saveRDS(FABLE %>% dplyr::filter(times >= 2020), here("Output", country, paste0(tag, "_FABLE.rds")))

# ns -> ns_int using your ns_map (built earlier)

results_DS$out.res <- format_to_id_map(results_DS, ns_map)

stopifnot(!anyNA(results_DS$out.res$ns))

downscaled_LUC <- results_DS$out.res

saveRDS(downscaled_LUC, here("Output", country, paste0(tag, "_downscaled_LUC.rds")))

```


# Downscaling - diagnostics & quality indicators

```{r restriction data, echo = FALSE, message=FALSE}

## Protected areas

#Almost no protected areas in India - will not use it

# Build a long mask of allowed = 0/1 per (ns, lu.to)
# pa_cells <- ProtectedAreas.df %>% 
#   pivot_longer(-c(id_c)) %>% 
#   mutate(code = stringr::str_remove(name, "X")) %>%
#   left_join(map_protectedareas %>% mutate(code = as.character(code))) %>% 
#   group_by(id_c) %>% 
#   #this is in km2 so I divide by 10 to get in 1000 ha - 1 km² = 0.1 thousand ha
#   summarise(value = sum(value,na.rm = T)/10) %>% 
#   left_join(grid %>% 
#               select(id_c, area) %>% 
#               #this is in km2 so I divide by 10 to get in 1000 ha - 1 km² = 0.1 thousand ha
#               mutate(area = area/10), by = "id_c") %>% 
#   mutate(p = 100 * value/area) #in %
# 
# restriction <- expand.grid(ns = unique(start_map$ns),
#                      lu.to = c("cropland","forest","newforest","otherland","pasture","urban"),
#                      stringsAsFactors = FALSE) %>%
#   mutate(value = 1L) %>%
#   mutate(value = ifelse(ns %in% pa_cells & lu.to %in% c("cropland","urban","pasture"), 0L, value))
# # 


#Use Griscom's global restoration map in t CO2/ 1000 ha
q75 <- quantile(rest_griscom.df[["p"]], prob=c(.75), type=1)

not_restoration_worthy <- rest_griscom.df %>% 
  filter(p < q75) %>% 
  merge(ns_map) %>% 
  select(ns)

rest_worthy <- setdiff(ns_map["ns"], not_restoration_worthy)

test_notworthy <- ns_map %>% 
  filter(ns %in% rest_worthy[[1]]) %>% 
  mutate(value = 1) %>% 
  rbind(ns_map %>% 
          filter(ns %in% not_restoration_worthy[[1]])
        %>% mutate(value = -1))%>%
  mutate(lu.from = "any",
         lu.to = "newforest",
         times = 2025) %>% 
  select(-ns) %>% 
  rename(ns = ns_int) %>% 
  select(ns, times, lu.from, lu.to, value)

results_DS_test <- results_DS
results_DS_test$out.res <- test_notworthy

plot_restricted <- LUC_plot(results_DS_test, rasterized_layer)

plot_restricted$LUC.plot
  
## /!\ /!\ /!\ /!\ /!\  
# restriction table needs to include exactly the same (lu.from, lu.to) transition pairs as in targets 
# (excluding diagonals where lu.from == lu.to). Using any subset/superset leads to errors or unexpected behavior.
## /!\ /!\ /!\ /!\ /!\ 

restriction <- data.frame(
  expand.grid(ns               = unique(start_map$ns),
              lu.from          = c("cropland", "forest", "otherland", "pasture", "urban"),
              lu.to            = c("cropland", "newforest", "otherland", "pasture", "urban", "forest"),
              stringsAsFactors = FALSE) %>%
    filter(!(lu.from == "forest" & lu.to == "newforest")) %>%
    filter(!(lu.from != "forest" & lu.to == "forest")) %>%
    filter(!(lu.from == lu.to)) %>%
    #allowed
    mutate(value = 0L) %>% 
    #not allowed
    mutate(value = ifelse((ns %in% not_restoration_worthy[[1]] & lu.to == "newforest"), 1L, value)) 
)


# # 
results_DS_rest <- downscale(
  targets     = FABLE %>% filter(times >= 2020),
  start.areas = country_start_areas,
  xmat        = X_long,
  betas       = pred_coeff_long,
  restrictions = restriction   # <-- if the API supports it
)

results_DS_rest$out.res <- format_to_id_map(results_DS_rest, ns_map)
# 
# saveRDS(results_DS, here("Output", country, paste0(tag, "_downscale.rds")))
# saveRDS(pred_coeff_long, here("Output", country, paste0(tag, "_mnl_betas.rds")))


saveRDS(results_DS_rest$out.res, here("Output", country, paste0(tag, "_downscaled_LUC_restriction_griscomq75.rds")))

```
```{r second set of restriction}

#Use Griscom's global restoration map in t CO2/ 1000 ha
q70_griscom <- quantile(rest_griscom.df[["p"]], prob=c(.05), type=1)

not_restoration_worthy_griscom <- rest_griscom.df %>% 
  filter(p < q70_griscom) %>% 
  merge(ns_map) %>% 
  select(ns)

#Use walker's unrealized potenial for increased storage of carbon on land
q70_walker <- quantile(rest_walker.df[["sum"]], prob=c(.05), type=1)

not_restoration_worthy_walker <- rest_walker.df %>% 
  filter(sum < q70_walker) %>% 
  merge(ns_map) %>% 
  select(ns)


rest_worthy <- setdiff(
    setdiff(ns_map["ns"], not_restoration_worthy_griscom),
    not_restoration_worthy_walker)

test_notworthy <- ns_map %>% 
  filter(ns %in% rest_worthy[[1]]) %>% 
  mutate(value = 1) %>% 
  rbind(ns_map %>% 
          filter(ns %in% not_restoration_worthy[[1]])
        %>% mutate(value = -1))%>%
  mutate(lu.from = "any",
         lu.to = "newforest",
         times = 2025) %>% 
  select(-ns) %>% 
  rename(ns = ns_int) %>% 
  select(ns, times, lu.from, lu.to, value)

results_DS_test <- results_DS
results_DS_test$out.res <- test_notworthy

plot_restricted <- LUC_plot(results_DS_test, rasterized_layer)

plot_restricted$LUC.plot

cells_notworthy <- ns_map %>% 
  filter(ns %in% unique(c(not_restoration_worthy_griscom[[1]],
                          not_restoration_worthy_walker[[1]]))) %>% 
  select(ns_int) %>% 
  rename(ns = ns_int) %>% 
  mutate(worthy = 0)

cells_worthy <- ns_map %>% 
  filter(!(ns %in% unique(c(not_restoration_worthy_griscom[[1]],
                          not_restoration_worthy_walker[[1]])))) %>% 
  select(ns_int) %>% 
  rename(ns = ns_int) %>% 
  mutate(worthy = 1)

# compute how much land can actually be afforested on these cells
area_restriction <- start_map_reproj %>% 
  left_join(ns_map) %>% 
  select(ns_int, lu.from, value) %>% 
  rename(ns = ns_int) %>% 
  left_join(rbind(cells_notworthy, cells_worthy)) %>% 
  filter(worthy == 1) %>% 
  filter(lu.from != "forest") %>% 
  filter(lu.from != "urban") %>% 
  pivot_wider(names_from = lu.from, values_from = value)

colSums(area_restriction)

# versus how much is planned on being afforested per lu.from type
area_target_restoration <- FABLE %>% 
  filter(lu.to == "newforest") %>% 
  pivot_wider(names_from = lu.from, values_from = value)

colSums(area_target_restoration %>% select(-c(times, lu.to)))

restriction_full <- data.frame(
  expand.grid(ns               = unique(start_map$ns),
              lu.from          = c("cropland", "forest", "otherland", "pasture", "urban"),
              lu.to            = c("cropland", "forest", "otherland", "pasture", "urban", "newforest"),
              stringsAsFactors = FALSE) %>%
    filter(!(lu.from == "forest" & lu.to == "newforest")) %>%
    filter(!(lu.from != "forest" & lu.to == "forest")) %>%
    filter(!(lu.from == lu.to)) %>%
    #allowed
    mutate(value = 0L) %>% 
    #not allowed
    mutate(value = ifelse((ns %in% (setdiff(ns_map["ns"], rest_worthy))$ns & lu.to == "newforest"), 1L, value)) 
)


# # Option A: If supported, pass to downscale()
results_DS_rest_full <- downscale(
  targets      = FABLE %>% filter(times >= 2020),
  start.areas  = country_start_areas,
  xmat         = X_long,
  betas        = pred_coeff_long,
  restrictions = restriction_full   # <-- if the API supports it
)

# #Error in error_restrictions[, ccc] : incorrect number of dimensions
# Error during wrapup: not that many frames on the stack
# Error: no more error handlers available (recursive errors?); invoking 'abort' restart


results_DS_rest_full$out.res <- format_to_id_map(results_DS_rest_full, ns_map)
saveRDS(results_DS_rest_full$out.res, here("Output", country, paste0(tag, "_downscaled_LUC_restriction_griscomq70_walkerq70.rds")))


```


## Aggregate consistency & diagonal zeros

This table compares the downscaled allocation with the national transition targets from the FABLE Calculator for each origin–destination land-use pair and year. For every (lu.from -> lu.to, time) it shows the allocated area, the target area, and a percentage error computed as 100 * (allocated − target) divided by the target (with a very small number added to avoid division by zero). 
How to interpret the results? Values should be very close to 0%. Systematically positive errors mean the optimizer allocates more than requested; negative errors mean less. Large discrepancies usually signal a mismatch in years, missing transitions in the targets, or binding restrictions that prevent exact matching.

```{r diagnostics & quality indicators - Aggregate consistency & diagonal zeros, echo=FALSE}

# Aggregate allocation vs targets

# helper to normalize times consistently
norm_year <- function(x) as.integer(as.character(x))

alloc_agg_fct <- function(res) {
  
  temp <- res$out.res
  
  alloc <- temp %>% 
    mutate(times = norm_year(times)) %>%
    group_by(times, lu.from, lu.to) %>%
    summarise(alloc = sum(value, na.rm = TRUE), .groups = "drop")
  
  return(alloc)
}

alloc_agg_norest <- alloc_agg_fct(results_DS)
alloc_agg_rest <- alloc_agg_fct(results_DS_rest)

targ_agg <- FABLE %>%
  mutate(times = norm_year(times)) %>%           # <- force same type as alloc_agg
  filter(times >= 2020) %>%
  group_by(times, lu.from, lu.to) %>%
  summarise(target = sum(value, na.rm = TRUE), .groups = "drop")

# now the join will match on identical types

consistency_fct <- function(alloc, targ){
  
  consistency <- full_join(alloc, targ, by = c("times","lu.from","lu.to")) %>%
    mutate(across(c(alloc, target), ~ replace_na(.x, 0)),
           pct_err = 100 * (alloc - target) / pmax(abs(target), 1e-9)) %>%
    filter(lu.from != lu.to)
  
  # if you filter by min/max year anywhere, make sure you're numeric:
  min_year <- min(consistency$times, na.rm = TRUE)
  print(
  knitr::kable(consistency %>% filter(times >= min_year)) 
  )
  writexl::write_xlsx(consistency %>% filter(times >= min_year), 
                      here("output", country, 
                           paste0(tag, "Diagnostic_allocation_vs_targets.xlsx")))
  
  
}
# consistency <- full_join(alloc_agg, targ_agg, by = c("times","lu.from","lu.to")) %>%
#   mutate(across(c(alloc, target), ~ replace_na(.x, 0)),
#          pct_err = 100 * (alloc - target) / pmax(abs(target), 1e-9)) %>%
#   filter(lu.from != lu.to)
# 
# # if you filter by min/max year anywhere, make sure you're numeric:
# min_year <- min(consistency$times, na.rm = TRUE)
# 
# knitr::kable(consistency %>% filter(times >= min_year)) 
# writexl::write_xlsx(consistency %>% filter(times >= min_year), 
#            here("output", country, 
#                 paste0(tag, "Diagnostic_allocation_vs_targets.xlsx")))

summary(consistency_fct(alloc_agg_norest, targ_agg)) 
summary(consistency_fct(alloc_agg_rest,   targ_agg))

```



<!-- ## Prior vs allocation alignment -->

<!-- This diagnostic compares the priors implied by the multinomial model (what the econometrics expects) with the actual allocation from the optimization (what downscalR did). -->

<!-- Top-20 percent capture: rank cells by the prior for a given transition and report the fraction of allocated area that falls into the top 20 percent prior cells.  -->
<!-- How to interpret the results? A random allocation would be about 0.20. Higher is better (for example, 0.50 is good, 0.70 is very strong). Values near 0.20 can mean the priors are flat or that strong constraints overrode them. -->

<!-- Divergence measures: summary numbers such as Kullback–Leibler or Jensen–Shannon divergence quantify how far the allocation distribution is from the prior distribution across cells.  -->
<!-- How to interpret the results? Lower is better; 0 means perfect agreement. -->

<!-- Rank correlation: a Spearman correlation between prior scores and allocated intensities. Interpretation: above 0.6 is strong monotone agreement; 0.3 to 0.6 is moderate; near 0 is weak. -->

<!-- Note: when priors are intentionally flat (because of limited data), low capture or high divergence does not necessarily indicate an error. -->

```{r compute true prior probabilities for QA (KL & Top-20 capture), echo = FALSE}

##Putting that on hold because does not work right now

# # ---- Prep (run once after you have country_df, pred_coeff_long, results_DS) ----
# # Use the already-scaled design for priors QC:
# X_wide_scaled <- country_df$xmat %>%
#   dplyr::filter(ks %in% keep_ks) %>%
#   dplyr::mutate(value = (value - mu[ks]) / sdv[ks]) %>%
#   tidyr::pivot_wider(names_from = "ks", values_from = "value", values_fill = 0) %>%
#   dplyr::arrange(ns)
# 
# X_ns  <- X_wide_scaled$ns
# X_mat <- as.matrix(dplyr::select(X_wide_scaled, -ns))
# features <- colnames(X_mat)
# classes  <- sort(unique(country_luc$lu.to))
# 
# # Numerically safe KL
# kl_div <- function(q, p, eps = 1e-12) {
#   q <- pmax(q, eps); p <- pmax(p, eps)
#   sum(q * log(q / p))
# }
# 
# # Optional: Jensen-Shannon divergence (symmetric, bounded)
# js_div <- function(p, q, eps = 1e-12) {
#   p <- pmax(p, eps); q <- pmax(q, eps)
#   m <- 0.5*(p+q)
#   0.5*kl_div(p, m, eps) + 0.5*kl_div(q, m, eps)
# }
# 
# # Reconstruct prior probabilities for an origin class (lu.from)
# prior_probs_for <- function(lu_from) {
#   # Make a (features x classes) coefficient matrix with zeros by default.
#   B <- matrix(0, nrow = length(features), ncol = length(classes),
#               dimnames = list(features, classes))
#   # Fill with estimated betas (long -> scatter into B)
#   cf <- pred_coeff_long %>% dplyr::filter(lu.from == !!lu_from)
#   ii <- match(cf$ks, features); jj <- match(cf$lu.to, classes)
#   keep <- which(!is.na(ii) & !is.na(jj))
#   if (length(keep)) B[cbind(ii[keep], jj[keep])] <- cf$value[keep]
#   # Baseline column (lu_from) remains zeros  - > correct for multinomial logit.
#   LP    <- X_mat %*% B
#   maxr  <- apply(LP, 1, max)
#   probs <- exp(LP - maxr); probs <- probs / rowSums(probs)
#   
#   list(ns = X_ns, classes = classes, probs = probs)
# }
# 
# # Compute KL & Top-k capture for one transition and one year
# compute_transition_qc <- function(lu_from, lu_to, year) {
#   pp  <- prior_probs_for(lu_from)
#   j   <- match(lu_to, pp$classes)
#   stopifnot(!is.na(j))
#   p   <- as.numeric(pp$probs[, j])       # prior per cell for (lu_from -> lu_to)
#   # Posterior allocation q from results (that year only)
#   q_tbl <- results_DS$out.res %>%
#     dplyr::filter(times == !!year, lu.from == !!lu_from, lu.to == !!lu_to) %>%
#     dplyr::group_by(ns) %>% dplyr::summarise(area = sum(value), .groups="drop")
#   q <- rep(0, length(pp$ns))
#   q[match(q_tbl$ns, pp$ns)] <- q_tbl$area
#   p <- p / pmax(sum(p), 1e-12)
#   q <- q / pmax(sum(q), 1e-12)
#   
#   # Metrics
#   Dkl <- kl_div(q, p)
#   Jsd <- js_div(p, q)
#   rho <- suppressWarnings(stats::cor(p, q, method = "spearman", use = "complete.obs"))
#   # Top-20/40
#   rk <- rank(-p, ties.method = "average")
#   n  <- length(rk)
#   top_capture <- function(frac) sum(q[rk <= frac*n])
#   data.frame(
#     year = year,
#     lu.from = lu_from,
#     lu.to = lu_to,
#     KL_q_p = Dkl,
#     JSD = Jsd,
#     Spearman = rho,
#     Top20 = top_capture(0.20)
#    
#   )
# }
# 
# yrs <- sort(unique(results_DS$out.res$times))
# base_year  <- min(yrs)
# target_yr  <- if (2050 %in% yrs) 2050 else max(yrs)
# 
# #full QC table for all transitions in a year
# pairs_eval <- FABLE %>% filter(times > 2020) %>% distinct(lu.from, lu.to)
# qc_table <- purrr::map_dfr(seq_len(nrow(pairs_eval)), function(i) {
#   with(pairs_eval[i,], compute_transition_qc(lu.from, lu.to, target_yr))
# })
# 
# 
# knitr::kable(qc_table %>% dplyr::arrange(KL_q_p))

# How to read the numbers (cheat-sheet)
# 
# Top20 ~ 0.20 -> Allocation is basically random with respect to the prior.
# Top20 >= 0.50 -> Strong alignment to prior hotspots.
# Top20 >= 0.70 -> Very strong; most mass sits where priors said it should.
# 
# KL (q||p): 0 means perfect match. There isn’t a universal “good” threshold because it depends on sparsity and number of cells, but:
# 
# 0.00–0.10: extremely close to priors
# 
# 0.10–0.50: modest deviations
# 
# >0.50: sizable deviation; check constraints or priors (may be flat/weak)
# 
# JSD (optional): symmetric and bounded in 
# [0,ln⁡2]
# [0,ln2] ~ 
# [0,0.693]
# [0,0.693]. Interprets like KL (lower is better), but easier to compare across cases.
# 
# Spearman: rank agreement between prior and allocation across cells.
# ~0 means weak monotone relation; >0.4 moderate; >0.6 strong.
# 
# Tip: If priors are very flat (little contrast), KL can be small and Top20 near 0.2 even if the map “looks okay”. That’s a prior issue—consider strengthening drivers or adding constraints.

```


# Land-use map

```{r land use map, echo = FALSE, message=FALSE, fig.align = "center", fig.width=8, fig.height=8, out.width='100%'}
# plot with consistent IDs on both sides
# plots per year land cover per land cover type 
LUC_plot_out <- LUC_plot(results_DS, rasterized_layer, label = "Area in 1000 ha per pixel")
plot_to_save <- LUC_plot_out$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(tag, "Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)

```

# Land-use change map

```{r land use change map, echo = FALSE, message = FALSE, fig.width=8, fig.height=8, out.width='100%'}

results_DS_LUC <- results_DS

# Create 'gain' and 'loss' datasets
results_DS_LUC$out.res <- loss_and_gains(results_DS_LUC) 

#Create the transition plot using LUC_plot
Transition_plot <- LUC_plot(
  results_DS_LUC, rasterized_layer, 
  color = "PRGn",
  label = "Area gains and losses in 1000 ha per pixel"
)

# Step 5: Create the plot with updated scale and theme adjustments
LUC_map <- Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))

#Save the plot to a .tiff file
tiff(
  here("Output", country, paste0(tag, "Land-use-change-map.tiff")),
  units = "in", height = 6, width = 10, res = 300
)
plot(LUC_map)

dev.off()


```


```{r, include = FALSE}
rm(LUC_map)
```

```{r 2020 map, echo = FALSE, message = FALSE}

results_DS_2050         <- results_DS
results_DS_2050$out.res <- results_DS_2050$out.res %>% 
  dplyr::filter(times %in% c(2020, 2050))

# Create the land-use change plot for 2050
LUC_plot_2050 <- LUC_plot(results_DS_2050, rasterized_layer)
plot_to_save  <- LUC_plot_2050$LUC.plot +
  ggplot2::theme(
    legend.key.width = ggplot2::unit(0.8, "cm"),
    panel.background = ggplot2::element_rect(fill = "lightgrey"))

plot_to_save

# Save the plot as a TIFF file

ggplot2::ggsave(
  filename = here("Output", country, paste0(tag, "2050_Land-use-map.tiff")), 
  plot = plot_to_save, 
  units = "in", 
  height = 6, width = 10, dpi = 300)
rm(plot_to_save)
```



```{r map cumulative change 2020-2050, echo =FALSE, message = FALSE}
cum_diff <- results_DS
cum_diff$out.res <- cum_difference(cum_diff)
            
LUC_plot_2020_2050 <- LUC_plot(cum_diff, rasterized_layer)
plot_to_save <- LUC_plot_2020_2050$LUC.plot +
  ggplot2::scale_fill_gradient2(low = "red",
                                  mid = "white",
                                  high = "blue")+
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm"),
                 panel.background = ggplot2::element_rect(fill = "lightgrey"))
plot_to_save      

ggplot2::ggsave(
  filename = here("Output", country, paste0(tag, "2020_2050_cumulative_final_Land-use_change-map.tiff")), 
  plot = plot_to_save, 
  units = "in", 
  height = 6, width = 10, dpi = 300)
rm(plot_to_save)

```

```{r transitions, echo =FALSE, message = FALSE}

### no restrictions ###

detailed_diff <- results_DS
detailed_diff$out.res <- pairwise_net_difference(results_DS, ns_map) 

Transition_plot_20_50 <- LUC_plot_cum_allLUC(
  detailed_diff, 
  rasterized_layer,
  label = "Transitions in % of pixel"
)

Transition_plot_20_50$LUC.plot
# View(Transition_plot_20_50)
# View(Transition_plot_20_50$plot.df)
# View(detailed_diff$out.res)

ggplot2::ggsave(
  filename = here("Output", country, paste0(tag, "2020_2050_norestriction_cumulative_pair_wise_Land-use_change-map.tiff")), 
  plot = Transition_plot_20_50$LUC.plot, 
  units = "in", 
  height = 6, width = 10, dpi = 300)
rm(plot_to_save)


### with restriction ####

detailed_diff_rest <- results_DS_rest
detailed_diff_rest$out.res <- pairwise_net_difference(results_DS_rest, ns_map)

Transition_plot_20_50_rest <- LUC_plot_cum_allLUC(
  detailed_diff_rest, 
  rasterized_layer,
  label = "Transitions in % of pixel"
)

Transition_plot_20_50_rest$LUC.plot

ggplot2::ggsave(
  filename = here("Output", country, paste0(tag, "2020_2050_restricted_cumulative_pair_wise_Land-use_change-map.tiff")), 
  plot = Transition_plot_20_50_rest$LUC.plot, 
  units = "in", 
  height = 6, width = 10, dpi = 300)
rm(plot_to_save)
```

```{r compare with restriction, echo =FALSE, message = FALSE}

results_several_pathways <- list(results_DS, results_DS_rest, results_DS_rest_full)
pathways_names <- c(paste0(params$pathway, "_norestriction"), paste0(params$pathway, "_WRI_restriction"), paste0(params$pathway, "_restrictions")  ) 

LUC_plot_pathways <- LUC_plot_compare_pathways(results_several_pathways, rasterized_layer, grid, pathways_names)
LUC_plot_pathways$LUC.plot

ggplot2::ggsave(
  filename = here("Output", country, paste0(tag, "2020_2050_per_pathway_cumulative_final_Land-use_change-map.tiff")), 
  plot = LUC_plot_pathways$LUC.plot, 
  units = "in", 
  height = 6, width = 10, dpi = 300)
#rm(LU_plot_pathways)
```


## Priors check

```{r priors_setup_scaled_X, echo=FALSE}
# Pre-req: these exist from the MNL/standardization step
stopifnot(exists("keep_ks"), exists("mu"), exists("sdv"))

# Build the SAME scaled design used for betas
X_wide_scaled <- country_df$xmat %>%
  dplyr::filter(ks %in% keep_ks) %>%
  dplyr::mutate(value = (value - mu[ks]) / sdv[ks]) %>%
  tidyr::pivot_wider(names_from = ks, values_from = value, values_fill = 0) %>%
  dplyr::arrange(ns)

X_mat  <- as.matrix(X_wide_scaled %>% dplyr::select(-ns))
classes <- sort(unique(country_luc$lu.to))  # destination set for priors
```

### Standardized coefficients


```{r priors_std_importance, echo=FALSE}

# Betas were estimated on standardized X, so betas are already "per 1 SD".
# Rank features by their overall effect size across destinations for a given origin.

std_importance <- function(lu_from, k_top = 3) {
  pred_coeff_long %>%
    dplyr::filter(lu.from == !!lu_from) %>%
    dplyr::group_by(ks) %>%
    dplyr::summarise(
      max_abs_beta = max(abs(value), na.rm = TRUE),
      l2_beta      = sqrt(sum(value^2, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(l2_beta)) %>%
    dplyr::slice_head(n = k_top)
}

std_importance_df <- data.frame()

land.from <- c("cropland","pasture","forest","otherland")
for(cur.from in land.from){
  std_importance_df <- dplyr::bind_rows(std_importance_df, cbind("land.from" = cur.from, std_importance(cur.from)))
}

writexl::write_xlsx(std_importance_df, 
           here("output", country, 
                paste0(tag, "Std_importance.xlsx")))

# Example table for one origin:
std_importance("otherland") %>% knitr::kable()

```

### Average Marginal Effects (AMEs) on probabilities

```{r Average Marginal Effects (AMEs) on probabilities, echo=FALSE}
# Build coefficient matrix for an origin class (baseline column = origin; zeros)
softmax <- function(M) { M <- sweep(M, 1, apply(M, 1, max), "-"); E <- exp(M); E/rowSums(E) }

coef_matrix_for <- function(lu_from) {
  B <- matrix(0, nrow = ncol(X_mat), ncol = length(classes),
              dimnames = list(colnames(X_mat), classes))
  cf <- pred_coeff_long %>% dplyr::filter(lu.from == !!lu_from)
  ii <- match(cf$ks, rownames(B)); jj <- match(cf$lu.to, colnames(B))
  keep <- which(!is.na(ii) & !is.na(jj))
  if (length(keep)) B[cbind(ii[keep], jj[keep])] <- cf$value[keep]
  B
}

average_marginal_effects <- function(lu_from) {
  B  <- coef_matrix_for(lu_from)
  P  <- softmax(X_mat %*% B)
  ame <- matrix(NA_real_, nrow = ncol(X_mat), ncol = ncol(B),
                dimnames = list(colnames(X_mat), colnames(B)))
  for (k in seq_len(ncol(X_mat))) {
    beta_k <- B[k, ]
    adj    <- P %*% beta_k
    dP     <- P * sweep(matrix(beta_k, nrow(P), ncol(P), byrow=TRUE), 1, adj, "-")
    ame[k, ] <- colMeans(dP)
  }
  ame
}


# Example: AMEs for one origin
AME <- average_marginal_effects("otherland")
# Rank features by overall AME magnitude across destinations
ame_rank <- data.frame(
  ks = rownames(AME),
  l1_ame = rowSums(abs(AME)),    # or l2_ame = sqrt(rowSums(AME^2))
  stringsAsFactors = FALSE
) |>
  dplyr::arrange(dplyr::desc(l1_ame))
head(ame_rank, 15)

```

### Permutation importance on priors

```{r priors permutation, echo = FALSE}
build_Ymat <- function(lu_from) {
  Y <- dplyr::filter(Yraw, lu.from == !!lu_from & Ts == 2015) |>
    tidyr::pivot_wider(names_from = lu.to, values_from = value, values_fill = 0) |>
    tibble::column_to_rownames("ns") |>
    dplyr::select(-lu.from, -Ts)
  Y <- Y[match(rownames(X_wide_scaled), rownames(Y)), , drop = FALSE]
  Y[is.na(Y)] <- 0
  as.matrix(Y)
}

logloss_prior <- function(lu_from, X = X_mat) {
  B  <- coef_matrix_for(lu_from)
  P  <- softmax(X %*% B) + 1e-12
  Y  <- build_Ymat(lu_from)
  -sum(Y * log(P))
}

perm_importance <- function(lu_from, features = colnames(X_mat), n_perm = 1) {
  base <- logloss_prior(lu_from)
  out <- lapply(features, function(f) {
    ll_inc <- replicate(n_perm, {
      Xp <- X_mat
      Xp[, f] <- sample(Xp[, f])
      logloss_prior(lu_from, X = Xp) - base
    })
    data.frame(ks = f, delta_logloss = mean(ll_inc))
  })
  dplyr::arrange(dplyr::bind_rows(out), dplyr::desc(delta_logloss))
}


# Example: top 15 most important features for origin "otherland"
#perm_importance("otherland") |> head(15)

```



