---
title: "Downscaling"
author: "Clara Douzal"
date: "2023-10-24"
output: pdf_document
params:
  variable1 : country
  variable2 : pathway 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include = FALSE}
library(geojsonio)
library(here)
library(raster)
library(dplyr)
library(tidyr)
library(downscalr)
library(stringr)
```


```{r temporary settings, include = FALSE}
#Please ensure that you have a folder named like the country variable you will input below in your "Output" folder. This is where your maps will be saved.

country = "IND"
pathway = "CurrentTrends"
# pathway = "NationalCommitments"
# pathway = "GlobalSustainability"

```

This report is generated for **`r country`** for the pathway **`r pathway`**.

# Checking input data

```{r data load, include=FALSE}

#Loading all the country specific data downloaded from Earth engine

#geospatial data

population <- geojson_read(here("Data", country, "Population2020.geojson"),  what = "sp")
population.df <- population@data %>% 
  select(-id) #%>% 
  #mutate(TotalPop = c(scale(TotalPop, scale = T)))

ProtectedAreas <- geojson_read(here("Data", country, "ProtectedAreas.geojson"),  what = "sp")
ProtectedAreas.df <- ProtectedAreas@data %>% 
  select(-id)

Travel <- geojson_read(here("Data", country, "TravelTime.geojson"),  what = "sp")
Travel.df <- Travel@data %>% 
  select(-id)

#this is in km2 
LandCover <- geojson_read(here("Data", country, "LandCoverHILDA2015.geojson"),  what = "sp")
LandCover.df <- LandCover@data %>% 
  select(-id)

#this is in km2
LandCoverChange <- geojson_read(here("Data", country, "LandCoverChangeHILDA2015_2019.geojson"),  what = "sp")
LandCoverChange.df <- LandCoverChange@data %>% 
   select(-id)

ForestManagement <- geojson_read(here("Data", country, "ForestManagement.geojson"),  what = "sp")
ForestManagement.df <- ForestManagement@data %>% 
  select(-id)

altitude <- geojson_read(here("Data", country, "Altitude.geojson"),  what = "sp")
altitude.df <- altitude@data %>% 
  select(-id)

slope <- geojson_read(here("Data", country, "Slope.geojson"),  what = "sp")
slope.df <- slope@data %>% 
  select(-id)
  

#mapping

map_HILDA <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA")

map_HILDA_LUC <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                        sheet = "HILDA_change")

map_Copernicus <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "Copernicus")

map_ForestMgmt <- readxl::read_excel(here("Data/mapping_code.xlsx"), 
                                sheet = "ForestManagement")
```

```{r create raster for maps, include = FALSE}
#raster to use for plots
ext <- extent(Travel)
res <- 0.5
raster_layer <- raster(ext, res = res)
rasterized_layer <- rasterize(Travel, raster_layer, field = "id")
```

Check in the table below that all land cover of interest and all possible transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with the same number in each cell (i.e. the number of pixels for your country's grid representation in 50km x 50km)

```{r changing LandCoverChange data structure , echo = FALSE, message = FALSE}
#value is the percentage of the cell that is going from land cover to another
country_luc <- LandCoverChange.df %>% 
  pivot_longer(-c(id_c), values_to = "AreaPerCode") %>% 
  #associate land cover code with right name
  mutate(name = stringr::str_remove(name, "X")) %>% 
  left_join(map_HILDA_LUC %>%  mutate(code = as.character(code)), by = c("name" = "code"))%>%
  #remove ocean and water
  filter(!(from %in% c("ocean", "water"))) %>% 
  mutate(AreaPerCode = as.numeric(AreaPerCode)) %>% 
  #group by cell
  group_by(id_c) %>% 
  #compute the total area in that cell
  mutate(TotalArea = sum(AreaPerCode, na.rm = T)) %>% 
  #compute the share of each land cover per cell
  mutate(value = AreaPerCode/TotalArea) %>%
  #set the right variable names to use in DownscalR package
  group_by(id_c, from, to) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(ns = id_c,
         lu.from = from,
         lu.to = to) %>% 
  mutate(Ts = 2015) %>% 
  dplyr::select(ns, lu.from, lu.to, Ts, value) %>% 
  mutate(value = ifelse(is.na(value), 0, value)) %>% 
  ####if lu.to is forest but coming from something else than lu.from = forest then it is new forest for FABLE design
  mutate(lu.to = ifelse(lu.from != "forest" & lu.to == "forest", "newforest", lu.to)) %>% 
  #add the land transition from any land cover to Forest and set it to zero because impossible in FABLE
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                                 each = length(unique(LandCoverChange.df$id_c))),
                   lu.to = "forest",
                   Ts = 2015,
                   value = 0)) %>% 
  #add the land transition from newforest to any other (but forest that has been added in previous step)
  # and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "newforest",
                   lu.to = rep(c("cropland", "newforest", "otherland", "pasture", "urban"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) %>% 
  #add land transition from forest to newforest and set it to zero so that the transition is in the matrix even though not in the calibration and won't be used
  rbind(data.frame(ns = unique(LandCoverChange.df$id_c),
                   lu.from = "forest",
                   lu.to = rep(c("newforest"), 
                               each = length(unique(LandCoverChange.df$id_c))),
                   Ts = 2015,
                   value = 0)) 

table(country_luc$lu.from, country_luc$lu.to)
  
```

# Harmonizing start map with FABLE Calculator baseline map

We need to harmonize the high-resolution baseline map used in the downscalR function with the aggregated land-use data from the FABLE Calculator. This ensures that the spatially explicit baseline map aligns with the aggregated land-use data in FABLE. Without this alignment, there could be discrepancies in the starting levels of each land-use class, making the future land-use transitions projected by the FABLE Calculator unrealistic when applied to the unadjusted high-resolution map.

```{r changing land cover structure to create start_map harmonized with FABLE aggregated land cover, echo = FALSE, message = FALSE}

#land use level per cell per land cover class in baseline year
start_map <- LandCover.df %>%
  # Pivot to long format and convert AreaPerCover to numeric
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_Copernicus and filter out "ocean" and "water"
  left_join(map_HILDA %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  filter(!LandCover %in% c("ocean", "water")) %>%
  # Group by id_c and LandCover, summarizing the total area per cover class
  group_by(id_c, LandCover) %>%
  #this is in km2 so I divide by 10 to get in 1000 ha
  summarise(value = sum(AreaPerCover, na.rm = TRUE)/10, .groups = 'drop') %>%
  # Rename columns for clarity
  rename(ns = id_c, lu.from = LandCover) %>%
  # Add rows for "newforest" with value = 0 for each unique id_c
  bind_rows(data.frame(ns = unique(LandCover.df$id_c), lu.from = "newforest", value = 0)) %>% 
  mutate(lu.from = ifelse(is.na(lu.from), "not relevant", lu.from))
  
start_map_harm <- start_map %>% 
  group_by(lu.from) %>% 
  summarise(value = sum(value, na.rm = T)) %>% 
  rename(lu = lu.from) %>% 
  mutate(lu = ifelse(is.na(lu), "not relevant", lu))

#FABLE Calculator aggregated land use level in baseline year
LC_targets <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = "Baseline") %>% 
  rename(lu = LandCover)

keep <- "from" # Specify which areas to keep during transition

# Calculate land-use change targets based on the start and target data
LUC_targets <- LU_to_LUC(start_map_harm, 
                              LC_targets, 
                              keep_areas = keep)

# Create a check data frame to validate the results
check_df <- rbind(NULL,
                  left_join(
                    start_map_harm %>% group_by(lu) %>% 
                      rename(lu.from = "lu") %>% 
                      reframe(value_start = sum(value)), 
                    LC_targets %>% group_by(lu) %>% 
                      rename(lu.to = "lu") %>% 
                      reframe(value_target = sum(value)), 
                    by = c("lu.from" = "lu.to")) %>%
                    left_join(LUC_targets %>% 
                                group_by(lu.to) %>% 
                                rename(value_new = "value") %>% 
                                reframe(value_new = sum(value_new)), 
                              by = c("lu.from" = "lu.to")) %>% 
                    select(lu.from, value_start, value_target, value_new) %>%
                    mutate(new_target_diff = value_new - value_target, 
                           value_rebalance = sum(value_start) * (value_target / sum(value_target)), 
                           rebalance_new_diff = value_rebalance - value_new))

### Calculate implied transition shares from LUC targets
CLC_FABLE_shares <- LUC_targets %>% 
  group_by(lu.from) %>% 
  mutate(value = value / sum(value)) %>% # Normalize values to get shares
  pivot_wider(id_cols = c(lu.from), names_from = "lu.to", values_from = "value") %>% 
  replace(is.na(.), 0) %>% # Replace NA with 0 for clarity
  pivot_longer(cols = !c(lu.from), names_to = "lu.to", values_to = "proj_share")

# Apply the transition shares to the high-resolution start map for downscaling
start_map_reproj <- left_join(start_map, CLC_FABLE_shares) %>% 
  mutate(value = ifelse(!is.na(lu.to), value * proj_share, value), # Adjust values based on shares
         lu.to = ifelse(is.na(lu.to), lu.from, lu.to)) %>% # Update land use type
  group_by(ns, lu.to) %>% 
  summarize(value = sum(value)) %>% # Sum values for each unique combination
  ungroup() %>% 
  drop_na() %>% # Remove rows with NA values
  rename(lu.from = "lu.to") # Rename for consistency


```
# Checking FABLE land use change projections to downscale

Check in the table below that in the land use changes projected in your calculator, all land cover of interest and all transitions are included in the land transition matrix.

-   The structure of the matrix should be such that you have the same rownames and colnames. Check that you have all land covers, by default it should be cropland, forest, new forest, otherland, pasture and urban.

-   The matrix should be filled with 10 in each cell (i.e. the number of year periods in the calculator).

-   The diagonal should be filled with zeros.

```{r FABLE data prep, echo = FALSE, message = FALSE}

FABLE <- readxl::read_excel(here("Data", country, "FABLE.xlsx"), sheet = pathway) %>%
  # In calculator, sheet 4_calc_land, 
  # table calc_landmatrix, columns LandCoverInit	YearStart	YearEnd	ToForest	ToOtherLand	ToCropland	ToPasture	ToUrban	ToNewForest
  
  # Remove YearStart and reshape to long format
  select(-YearStart) %>%
  pivot_longer(-c(YearEnd, LandCoverInit), values_to = "value", names_to = "lu.to") %>%
  # Clean and adjust land cover names
  mutate(
    lu.from = tolower(LandCoverInit),
    lu.to = tolower(str_remove(lu.to, "To")),
    times = YearEnd
  ) %>%
  # Select necessary columns and filter out unchanged land covers
  select(times, lu.from, lu.to, value) %>%
  filter(lu.from != lu.to)
  
table(FABLE$lu.from, FABLE$lu.to)

```


```{r create Xmat matric of explanatory variables, echo = FALSE, message = FALSE}

Xmat_LU <- start_map_reproj %>% 
  pivot_wider(names_from = "lu.from", values_from = "value", values_fill = 0) %>% 
  rename(id_c = ns)

#ForestManagement
Xmat_ForestManagement <- ForestManagement.df %>%
  # Pivot longer to convert cover columns to rows
  pivot_longer(-id_c, values_to = "AreaPerCover") %>%
  mutate(AreaPerCover = as.numeric(AreaPerCover),
         name = str_remove(name, "X")) %>%
  # Join with map_ForestMgmt to get the corresponding forest management types
  left_join(map_ForestMgmt %>% mutate(code = as.character(code)), by = c("name" = "code")) %>%
  # Group by id_c and type to summarize total area per cover class per pixel
  group_by(id_c, type) %>%
  summarise(value = sum(AreaPerCover, na.rm = TRUE), .groups = 'drop') %>%
  # Pivot wider to create columns for each forest management type
  pivot_wider(names_from = type, values_from = value, values_fill = 0) 


#Mean Travel time
Xmat_Travel <- Travel.df %>% select(id_c, MeanTravelTime) %>% 
  mutate(log_MeanTravelTime = log(MeanTravelTime)) %>% 
  select(-MeanTravelTime)
#summary(Xmat_Travel)

#Population  
Xmat_pop <- population.df %>% select(id_c, TotalPop) %>% 
  mutate(log_pop = log(TotalPop)) %>% 
  select(-TotalPop)
#summary(Xmat_pop)  
  
#Altitude  
Xmat_alt <- altitude.df %>% select(id_c, MeanAltitude) %>% 
  mutate(log_altitude = log(MeanAltitude)) %>% 
  select(-MeanAltitude)

mode_thermal <- slope.df %>%
    select(id_c, thermal) %>%
    group_by(thermal) %>%          # Group by the "thermal" column
    summarize(count = n()) %>%     # Count the occurrences of each unique value in "thermal"
    arrange(desc(count)) %>%       # Sort by the count in descending order
    slice(1)    

mode_zone <- slope.df %>%
    select(id_c, zone) %>%
    group_by(zone) %>%          # Group by the "thermal" column
    summarize(count = n()) %>%     # Count the occurrences of each unique value in "thermal"
    arrange(desc(count)) %>%       # Sort by the count in descending order
    slice(1)    

#Slope and data from GAEZ on zone and thermal 
Xmat_slope <- slope.df %>%
  select(id_c, MeanSlope, thermal, zone) %>%
  # Create binary columns for each thermal category
  pivot_wider(id_cols = c(id_c, MeanSlope), 
              names_from = thermal, 
              values_from = thermal, 
              names_prefix = "thermal_", 
              values_fn = list(thermal = ~ 1), 
              values_fill = list(thermal = 0)) %>%
  # Retain MeanSlope before the second pivot for zone
  left_join(
    slope.df %>%
      select(id_c, zone) %>%
      pivot_wider(id_cols = id_c, 
                  names_from = zone, 
                  values_from = zone, 
                  names_prefix = "zone_", 
                  values_fn = list(zone = ~ 1), 
                  values_fill = list(zone = 0)), 
    by = "id_c"
  ) %>%
  mutate(log_slope = log(MeanSlope)) %>%
  select(-c(MeanSlope, 
            paste("thermal", mode_thermal$thermal, sep = "_"),
            paste("zone", mode_zone$zone, sep = "_")))
  


Xmat <- Xmat_LU %>% 
  #select(-forest_share) %>% 
  mutate(Intercept = 1) %>% 
  left_join(Xmat_Travel) %>% 
  left_join(Xmat_pop) %>% 
  left_join(Xmat_ForestManagement) %>% 
  left_join(Xmat_alt) %>% 
  left_join(Xmat_slope) %>% 
 # select(-c(cropland, otherland, urban, pasture, forest)) %>% #, SRP_share)) %>% 
  #pivot back longer to fit the DownscalR package structure
  pivot_longer(-id_c, values_to = "value", names_to = "ks") %>% 
  rename(ns = id_c)

```


```{r create df used for priors relative to the country, echo = FALSE, message=FALSE}
#find where I have NAs in the Xmat matrix to remove those cells else the MNL cannot run
rm_na <- Xmat$ns[which(is.na(Xmat), arr.ind = T)[,1]]

country_df <- list(xmat = tibble(Xmat %>% 
                                   filter(!(ns %in% rm_na))), 
                   lu_levels = tibble(start_map_reproj %>% 
                                        filter(!(ns %in% rm_na))))
```


# Estimate MNL model

```{r prep data, echo = FALSE, message=FALSE, include = FALSE}

results_MNL <- data.frame()
pred_coeff_long <- data.frame()
country_start_areas <- data.frame()
X_long <- data.frame()

Yraw <- country_luc %>% 
  filter(!(ns %in% rm_na))

for(cur_LU.from in c("urban", "cropland", "pasture", "forest", "otherland", "newforest")){
  
 Y <- dplyr::filter(Yraw,lu.from == cur_LU.from & Ts == 2015) %>% #changes 2018 to 2019
  tidyr::pivot_wider(names_from = lu.to) %>%
  tibble::column_to_rownames(var = "ns") %>%
  dplyr::select(-c(lu.from, Ts))
 
 X <- country_df$xmat %>% 
    tidyr::pivot_wider(names_from = "ks") %>%
    dplyr::arrange(match(ns,Y$ns)) %>%
    tibble::column_to_rownames(var = "ns") 
  
  baseline <- which(colnames(Y) == cur_LU.from)

  # Decreased niter = 100 and nburn = 50 for faster computation 
  temp_results_MNL <- mnlogit(as.matrix(X), as.matrix(Y), baseline, niter = 100, nburn = 50, A0 = 10^4)
  
  pred_coeff <- as.data.frame(apply(temp_results_MNL$postb[,-baseline,],c(1,2),mean))
  
  pred_coeff_long <- pred_coeff_long %>%
    rbind(
    pred_coeff %>% tibble::rownames_to_column(var="ks") %>%
  tidyr::pivot_longer(!ks, names_to = "lu.to", values_to="value") %>%
  dplyr::arrange(lu.to) %>%
  tibble::add_column(lu.from=cur_LU.from, .after="ks")
    )
  
  ## starting area should provide information on area of each (higher resolution) grid cell, i.e ns
country_start_areas <- country_start_areas %>%
  rbind(
    dplyr::filter(country_df$lu_levels, lu.from == cur_LU.from) %>% mutate(times = 2020)
  )

}
```

# Downscaling

```{r restriction data, echo = FALSE, message=FALSE}

## Protected areas

```



```{r, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}


X_long <-  X %>% tibble::rownames_to_column(var="ns") %>%
  tidyr::pivot_longer(!ns, names_to="ks", values_to = "value")

## Downscaling computation and results

results_DS <- downscale(targets = FABLE %>% filter(times > 2020), 
                        start.areas =  country_start_areas, 
                        xmat = X_long, 
                        betas = pred_coeff_long)

downscaled_LUC <- results_DS$out.res

```

# Land-use map

```{r, echo = FALSE, message=FALSE}
LUC_plot <- LUC_plot(results_DS, rasterized_layer)
plot_to_save <- LUC_plot$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)

```

# Land-use change map

```{r, echo = FALSE, message = FALSE}

results_DS_LUC <- results_DS

# Create 'gain' and 'loss' datasets
results_DS_LUC$out.res <- results_DS_LUC$out.res %>%
  filter(lu.from != lu.to) %>%
  # Group by `lu.to`, `ns`, and `times` for gains
  group_by(lu.to, ns, times) %>%
  summarise(gain = sum(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(Type = "gain", lu = lu.to) %>%
  
  # Bind the losses dataset by grouping by `lu.from`
  bind_rows(
    results_DS_LUC$out.res %>%
      filter(lu.from != lu.to) %>%
      group_by(lu.from, ns, times) %>%
      summarise(loss = -sum(value, na.rm = TRUE), .groups = "drop") %>%
      mutate(Type = "loss", lu = lu.from)
  ) %>%
  
  # Summarize the final dataset grouping by land-use (`lu`), time, and pixel (`ns`)
  group_by(lu, times, ns) %>%
  summarise(value = sum(gain, loss, na.rm = TRUE), .groups = "drop") %>%
  rename(lu.to = lu)

#Define plot aesthetics
color = "PRGn"
label = "Area gains and losses in ha per pixel"

#Create the transition plot using LUC_plot
Transition_plot <- LUC_plot(
  results_DS_LUC, rasterized_layer, 
  color = color,
  label = label
)

#Define the limit for color scaling
limit <- max(abs(Transition_plot[["plot.df"]]$value)) * c(-1, 1)

# Step 5: Create the plot with updated scale and theme adjustments
Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))

#Save the plot to a .tiff file
tiff(
  here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "Land-use-change-map.tiff")),
  units = "in", height = 6, width = 10, res = 300
)
plot(Transition_plot$LUC.plot + 
  ggplot2::scale_fill_gradient2() +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.5, "cm"))
)
dev.off()



```

```{r 2020 map, echo = FALSE, message = FALSE}

results_DS_2050 <- results_DS
results_DS_2050$out.res <- results_DS_2050$out.res %>% 
  dplyr::filter(times %in% c(2015, 2050))

# Create the land-use change plot for 2050
LUC_plot_2050 <- LUC_plot(results_DS_2050, rasterized_layer)
plot_to_save <- LUC_plot_2050$LUC.plot +
  ggplot2::theme(legend.key.width = ggplot2::unit(0.8, "cm"))
plot_to_save

# Save the plot as a TIFF file
output_file <- here("Output", country, paste0(format(Sys.Date(), "%y%m%d"), "_", pathway, "_", "2050_Land-use-map.tiff"))

ggplot2::ggsave(filename = output_file, plot = plot_to_save, units = "in", height = 6, width = 10, dpi = 300)
rm(plot_to_save)
```

